---
title: "chi-squared distribution"
author: "David Gerard"
date: today
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
            fig.width = 4, 
            fig.height = 3, 
            fig.align = "center")
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

## Learning Objectives

- Work with $\chi^2$-distribution
- Understand $\chi^2$-distribution

# $\chi^2$-distribution

- $\chi^2$ with different degrees of freedom

  ```{r, echo = FALSE, message=FALSE, warning=FALSE}
  library(tidyverse)
  library(ggthemes)
  x <- seq(0,20, length = 200)
  data.frame(df = as.factor(c(rep(2, length(x)), rep(3, length(x)), rep(4, length(x)))),
             x = c(x, x, x),
             y = c(dchisq(x = x, df = 2),
                   dchisq(x = x, df = 3),
                   dchisq(x = x, df = 4))) ->
    dfdat
  ggplot(dfdat, mapping = aes(x = x, y = y, color = df, lty = df)) +
    geom_line() +
    scale_color_colorblind() +
    ylab("f(x)")
  ```

# Working with $t$ Distribution in R:

- Density Function

  ```{r}
  dchisq(x = 3, df = 3)
  ```
  
  ```{r, echo = FALSE}
  library(ggplot2)
  x <- seq(0, 12, length = 100)
  y <- dchisq(x, df = 3)
  data.frame(x = x, y = y) |>
    ggplot(aes(x = x, y = y)) +
    geom_line() +
    ylab("f(x)") +
    geom_segment(dat = data.frame(x = 3, xend = 3, y = 0, yend = dchisq(x = 3, df = 3)), aes(x = x, y = y, xend = xend, yend = yend), lty = 2, col = 2)
  ```

- Random Generation

  ```{r}
  samp <- rchisq(n = 1000, df = 3)
  head(samp)
  ```
  
  ```{r, echo = FALSE}
  ggplot(tibble(samp), aes(x = samp)) +
    geom_histogram(fill = "white", color = "black", bins = 20)
  ```


- Cumulative Distribution Function

  ```{r}
  pchisq(q = 3, df = 3)
  ```
  
  ```{r, echo = FALSE}
  x <- seq(0, 12, length = 500)
  y <- dchisq(x,  df = 3)
  polydf <- data.frame(x = c(min(x), x[x < 3], 3, 0), 
                       y = c(0, y[x < 3], 0, 0))
  ggplot(tibble(x, y), aes(x = x, y = y)) +
    geom_line() +
    ylab("f(x)") +
    geom_polygon(data = polydf, mapping = aes(x = x, y = y), 
                 fill = "red", alpha = 3/4)
  ```

- Quantile Function: Rosner denotes the following $t_{2,0.9082}$

  ```{r}
  qchisq(p = 0.6083, df = 3)
  ```
  
  ```{r, echo = FALSE}
  x <- seq(0, 12, length = 500)
  y <- dchisq(x,  df = 3)
  polydf <- data.frame(x = c(min(x), x[x < 3], 3, 0), 
                       y = c(0, y[x < 3], 0, 0))
  ggplot(tibble(x, y), aes(x = x, y = y)) +
    geom_line() +
    ylab("f(x)") +
    geom_polygon(data = polydf, mapping = aes(x = x, y = y), 
                 fill = "red", alpha = 3/4) +
    annotate(geom = "text", x = 1.5, y = 0.1, label = "0.6083", color = "white")
  ```

::: {.panel-tabset}
## Exercise
In the Rosner notation, what does $\chi^2_{20, 0.95}$ mean? Calculate it.

## Solution
It is the 0.95 quantile of a chi-squared distribution with 20 degrees of freedom
```{r}
qchisq(p = 0.95, df = 20)
```
:::

