---
title: "t-distribution"
author: "David Gerard"
date: today
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
            fig.width = 4, 
            fig.height = 3, 
            fig.align = "center")
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

## Learning Objectives

-   Work with $t$-distribution
-   Understand $t$-distribution

# $t$-distribution

-   $t$ vs Normal(0, 1)

    ```{r}
    #| echo: false
    #| message: false
    #| warning: false
    #| fig-alt: "Density plots of t-distributions with different degrees of freedom."
    library(tidyverse)
    library(ggthemes)
    x <- seq(-4, 4, length = 200)
    data.frame(df = as.factor(c(rep(1, length(x)), rep(5, length(x)), rep(Inf, length(x)))),
               x = c(x, x, x),
               y = c(dt(x = x, df = 1),
                     dt(x = x, df = 5),
                     dt(x = x, df = Inf))) ->
      dfdat
    ggplot(dfdat, mapping = aes(x = x, y = y, color = df, lty = df)) +
      geom_line() +
      scale_color_colorblind() +
      ylab("f(x)")
    ```

-   A lot more extreme values in a t

    ```{r}
    #| echo: false
    #| fig-alt: "Histogram from normal distribution."
    set.seed(1)
    tibble(samp = rnorm(1000)) |>
      ggplot(aes(x = samp)) +
      geom_histogram(bins = 30, fill = "white", color = "black") +
      ggtitle("Normal") +
      xlab("Sample Values")
    ```
  
    ```{r}
    #| echo: false
    #| fig-alt: "Histogram from t-distribution with 2 degrees of freedom."
    set.seed(1)
    tibble(samp = rt(1000, df = 2)) |>
      ggplot(aes(x = samp)) +
      geom_histogram(bins = 30, fill = "white", color = "black") +
      ggtitle("t with 2 df") +
      xlab("Sample Values")
    ```

# Working with $t$ Distribution in R:

-   Density Function

    ```{r}
    dt(x = -6, df = 2)
    ```
    
    ```{r}
    #| echo: false
    #| fig-alt: "dt() is height of density curve."
    library(ggplot2)
    x <- seq(-6, 6, length = 100)
    y <- dt(x, df = 1)
    qplot(x, y, geom = "line", ylab = "f(x)") +
      geom_segment(dat = data.frame(x = 2, xend = 2, y = 0, yend = dt(x = 2, df = 2)), aes(x = x, y = y, xend = xend, yend = yend), lty = 2, col = 2)
    ```

-   Random Generation

    ```{r}
    samp <- rt(n = 1000, df = 2)
    head(samp)
    ```
    
    ```{r}
    #| echo: false
    #| fig-alt: "rt() gives you random samples."
    qplot(samp, geom = "histogram", fill = I("white"), color = I("black"), bins = 20)
    ```


-   Cumulative Distribution Function

    ```{r}
    pt(q = 2, df = 2)
    ```
    
    ```{r}
    #| echo: false
    #| fig-alt: "pt() gives you probability of being less than a value."
    x <- seq(-6, 6, length = 500)
    y <- dt(x,  df = 2)
    polydf <- data.frame(x = c(min(x), x[x < 2], 2, 0), 
                         y = c(0, y[x < 2], 0, 0))
    qplot(x, y, geom = "line", ylab = "f(x)") +
      geom_polygon(data = polydf, mapping = aes(x = x, y = y), 
                   fill = "red", alpha = 3/4)
    ```

-   Quantile Function: Rosner denotes the following $t_{2,0.9082}$

    ```{r}
    qt(p = 0.9082, df = 2)
    ```
    
    ```{r}
    #| echo: false
    #| fig-alt: "qt() is inverse of pt()"
    x <- seq(-6, 6, length = 500)
    y <- dt(x, df = 2)
    polydf <- data.frame(x = c(min(x), x[x < 2], 2, 0), 
                         y = c(0, y[x < 2], 0, 0))
    qplot(x, y, geom = "line", ylab = "f(x)") +
      geom_polygon(data = polydf, mapping = aes(x = x, y = y), 
                   fill = "red", alpha = 3/4) +
      annotate(geom = "text", x = 0, y = 0.1, label = "0.9082", color = "white")
    ```

::: {.panel-tabset}
## Exercise
In the Rosner notation, what does $t_{20, 0.95}$ mean? Calculate it.

## Solution
It's the 0.95 quantile of a $t$-distribution with 20 degrees of freedom
```{r}
qt(p = 0.95, df = 20)
```
:::
