---
title: "Chapter 4 Practice Problems"
author: "David Gerard"
date: today
---

These practice problems mostly come from Rosner's publicly available study sheet at the books [companion website](https://www.cengage.com/cgi-wadsworth/course_products_wp.pl?fid=M20b&product_isbn_issn=9781305268920). The solutions are my own, since we differ slightly on what we are looking for in the solutions.

```{r}
#| echo: false
#| eval: true
#| message: false
library(tidyverse)
```


## Basic Discrete Probability

::: {.panel-tabset} 
## 4.1 
Evaluate the number of ways of selecting 4 objects out of 10 if the order of selection matters. What term is used to denote this quantity?

## Solution
The book's notation is $_{10}P_4$. This is the number of permutations of 10 things taken 4 times. This is $10 \times 9 \times 8 \times 7 = 5040$.
:::

::: {.panel-tabset} 
## 4.2 
Evaluate the number of ways of selecting 4 objects out of 10 if the order of selection does not matter. What term is used to denote this quantity?

## Solution
This is the number of combination sof 10 things taken 4 times. The book notation is $_{10}C_4$, but everyone else uses $\binom{10}{4}$. 
$$
\binom{10}{4} = \frac{10 \times 9 \times 8 \times 7}{4 \times 3 \times 2 \times 1} = 210
$$
Checking in R, we get:
```{r}
choose(10, 4)
```

:::

::: {.panel-tabset} 
## 4.3 
Suppose that the probability that a person will develop hypertension over a lifetime is 20%. What is the probability distribution of the number of hypertensives over a lifetime among 20 students graduating from the same high school class?

## Solution
Binomial(20, 0.2)
:::

::: {.panel-tabset} 
## 4.4 
What is the probability that exactly 4 people out of 50 aged 60–64 will die after receiving the flu vaccine if the probability that 1 person will die is .028?

## Solution
Let $X$ be the number that die out of 50. Then $X \sim \mathrm{Binom}(50, 0.028)$. We calculate $P(X = 4)$ via
```{r}
dbinom(x = 4, size = 50, prob = 0.028)
```

:::

::: {.panel-tabset} 
## 4.5 
What is the probability that at least 4 people will die after receiving the vaccine?

## Solution
```{r}
1 - pbinom(q = 3, size = 50, prob = 0.028)
```
:::

::: {.panel-tabset} 
## 4.6 
What is the expected number of deaths following the flu vaccine?

## Solution
$np$ = 50 $\times$ 0.028 = 1.4.
:::

::: {.panel-tabset} 
## 4.7 
What is the standard deviation of the number of deaths following the flu vaccine?

## Solution
$$
\sqrt{np(1-p)} = \sqrt{50 \times 0.028 \times (1 - 0.028)}
$$
Evalulating in R, we get
```{r}
sqrt(50 * 0.028 * (1 - 0.028))
```

:::


## Health-Services Administration

The in-hospital mortality rate for 16 clinical conditions at 981 hospitals was recently reported. It was reported that in-hospital mortality was 10.5% for coronary-bypass surgery and 5.0% for total hip replacement. Suppose an institution changes from an academic institution to a private for-profit institution. They find that after the change, of the first 20 patients receiving coronary-bypass surgery, 5 die, while of 20 patients receiving total hip replacement, 4 die.

::: {.panel-tabset} 
## 4.8 
What is the probability that of 20 patients receiving coronary-bypass surgery, exactly 5 will die in-hospital, if this hospital is representative of the total pool of 981 hospitals?

## Solution
Let $X$ be the number among the coronary-bypass patients that die out of 20. Then $X \sim \mathrm{Binom}(20, 0.105)$. We want $P(X = 5)$
```{r}
dbinom(x = 5, size = 20, prob = 0.105)
```

:::

::: {.panel-tabset} 
## 4.9 
What is the probability of at least 5 deaths among the coronary-bypass patients?

## Solution
We want $P(X \geq 5) = 1 - P(X \leq 4)$
```{r}
1 - pbinom(q = 4, size = 20, prob = 0.105)
```

:::

::: {.panel-tabset} 
## 4.10
What is the probability of no more than 5 deaths among the coronary-bypass patients?

## Solution
We want $P(X \leq 5)$
```{r}
pbinom(q = 5, size = 20, prob = 0.105)
```

:::

::: {.panel-tabset} 
## 4.11 
What is the probability that exactly 4 will die among the hip-replacement patients?

## Solution
Let $Y$ be the number that die among the hip-replacement patients out of 20. Then $Y \sim \mathrm{Binom}(20, 0.05)$. We want $P(X = 4)$
```{r}
dbinom(x = 4, size = 20, prob = 0.05)
```

:::

::: {.panel-tabset} 
## 4.12 
What is the probability that at least 4 will die among the hip-replacement patients?

## Solution
We want $P(X \geq 4) = 1 - P(X \leq 3)$
```{r}
1 - pbinom(q = 3, size = 20, prob = 0.05)
```

:::

::: {.panel-tabset} 
## 4.13 
What is the probability of 4 or fewer deaths among the hip-replacement patients?

## Solution
We want $P(X \leq 4)$
```{r}
pbinom(q = 4, size = 20, prob = 0.05)
```

:::

::: {.panel-tabset} 
## 4.14
Can you draw any conclusions based on the results in Problems 4.8–4.13 regarding any effects of the change in hospital administration on in-hospital mortality rates?

## Solution
It would be very unlikely that this many patients would die if this hospital was representative of the 981 hospitals. Assuming that it was representative, we would only have a 5.2% chance of seeing as many or more deaths among the coronary-artery patients, and we would only have a 1.6% chance of seeing as many or more hip-deaths.
:::

## Cardiovascular Disease 1

The rate of myocardial infarction (MI) in 50–59-year-old, disease-free women is approximately 2 per 1000 per year or 10 per 1000 over 5 years. Suppose that 3 MI’s are reported over 5 years among 1000 postmenopausal women initially disease free who have been taking postmenopausal hormones.

::: {.panel-tabset} 
## 4.15
Use the binomial distribution to see if this experience represents an unusually small number of events based on the overall rate.

## Solution
Let $X$ be the number of MI's out of 1000 over the 5 years. The probability of MI over that time is 10/1000 = 0.01. Then we have that $X \sim \mathrm{Binom}(1000, 0.01)$. We want $P(X \leq 3)$, because this probability tells us if $X$ is unusually small:
```{r}
pbinom(q = 3, size = 1000, prob = 0.01)
```
So it is unusually small. We would only expect to see this 1% of the time.
:::

::: {.panel-tabset} 
## 4.16
Answer Problem 4.15 using the Poisson approximation to the binomial distribution.

## Solution
$X \approx \mathrm{Pois}(1000 \times 0.01) = \mathrm{Pois}(10)$. Calculation $P(X\leq 3)$ we get:
```{r}
ppois(q = 3, lambda = 10)
```

:::

::: {.panel-tabset} 
## 4.17 
Compare your answers in Problems 4.15 and 4.16.

## Solution
They are close.
:::

## Cardiovascular Disease 2

::: {.panel-tabset} 
## 4.18
An accepted hypothesis in the etiology of heart disease is that aspirin intake of 325 mg per day reduces subsequent cardiovascular mortality in men with a prior heart attack. Suppose that in a pilot study of 50 women who received 1 tablet per day (325 mg), only 2 die over a 3-year period from cardiovascular disease. How likely is it that not more than 2 women will die if the underlying 3-year mortality rate is 10% in such women?

## Solution
Let $X$ be the number of women that die out of 50. Then $X \sim \mathrm{Binom}(50, 0.1)$. We want to calculate $P(X \leq 2)$:
```{r}
pbinom(q = 2, size = 50, prob = 0.1)
```

:::

## Pediatrics

A hospital administrator wants to construct a special-care nursery for low-birthweight infants (≤ 2500 g) and wants to have some idea as to the number of beds she should allocate to the nursery. She is willing to assume that the recovery period of each baby is exactly 4 days and thus is interested in the expected number of premature births over the period.

::: {.panel-tabset} 
## 4.19 
If the number of premature births in any 4-day period is binomially distributed with parameters n = 25 and p = .1, then find the probability of 0, 1, 2, ..., 7 premature births over this period.

## Solution
Let $X$ be the number of premature births out of 25. Then $X \sim \mathrm{Binom}(25, 0.1)$. We want $P(X = 0), P(X = 1),\ldots,P(X = 7)$. We can do this in R via:
```{r}
pvec <- dbinom(x = 0:7, size = 25, prob = 0.1)
pvec
```

:::

::: {.panel-tabset} 
## 4.20 
The administrator wishes to allocate $x$ beds where the probability of having more than $x$ premature births over a 4-day period is less than 5%. What is the smallest value of $x$ that satisfies this criterion?

## Solution
This is the same thing as finding the smallest value $x$ such that $P(X \leq x) > 0.95$. This is exactly the 0.95 quantile from the binomial distribution.
```{r}
qbinom(p = 0.95, size = 25, prob = 0.1)
```
So, 5 beds.

We can confirm by seeing if  $P(X \geq 5)$ is less than 0.05 and $P(X \geq 4)$ is greater than 0.05.
```{r}
1 - pbinom(q = 5, size = 25, prob = 0.1)
1 - pbinom(q = 4, size = 25, prob = 0.1)
```
:::

::: {.panel-tabset} 
## 4.21 
Answer Problem 4.20 for 1%.

## Solution
Using the same logic, we want the 0.99 quantile of a $\mathrm{Binom}(25, 0.1)$ distribution
```{r}
qbinom(p = 0.99, size = 25, prob = 0.1)
```
So, six beds. Verifying, we get
```{r}
1 - pbinom(q = 6, size = 25, prob = 0.1)
1 - pbinom(q = 5, size = 25, prob = 0.1)
```
:::

## Cancer 1

The incidence rate of malignant melanoma is suspected to be increasing over time. To document this increase, a questionnaire was mailed to 100,000 U.S. nurses in 1976 and 1978, asking about any current or previous tumors. Thirty new cases of malignant melanoma were found to have developed over the 2-year period among women with no previous cancers in 1976.

::: {.panel-tabset} 
## 4.22 
If the annual incidence rate from previous cancer-registry data is 10 per 100,000, then what is the expected number of new cases over 2 years?

## Solution
$$
100000 \times \frac{10}{100000} \times 2 = 20
$$
:::

::: {.panel-tabset} 
## 4.23 
Are the preceding results consistent or inconsistent with the cancer-registry data? Specifically, what is the probability of observing at least 30 new cases over a 2-year period if the cancer-registry incidence rate is correct?

## Solution
The exact distribution is a $\mathrm{Binom}(100000, 20/100000)$. We want $P(X \geq 30)$
```{r}
1 - pbinom(q = 29, size = 100000, prob = 20/100000)
```
You can also use a $\mathrm{Pois}(20)$
```{r}
1 - ppois(q = 29, lambda = 20)
```
:::

## Accident Epidemiology
Suppose the annual number of traffic fatalities at a given intersection follows a Poisson distribution with parameter µ = 10 .

::: {.panel-tabset} 
## 4.24 
What is the probability of observing exactly 10 traffic fatalities in 1992?

## Solution
```{r}
dpois(x = 10, lambda = 10)
```

:::

::: {.panel-tabset} 
## 4.25
What is the probability of observing exactly 25 traffic fatalities over the 2-year period from January 1, 1990, to December 31, 1991?

## Solution
Let $X$ be the number of fatalities over 2 years. Then $X \sim \mathrm{Pois}(20)$. We want $P(X = 25)$
```{r}
dpois(x = 25, lambda = 20)
```

:::

::: {.panel-tabset} 
## 4.26 
Suppose that the traffic intersection is redesigned with better lighting, and 12 traffic fatalities are observed over the next 2 years. Is this rate a meaningful improvement over the previous rate of traffic fatalities?

## Solution
Let's calculate $P(X \leq 12)$ when $X \sim \mathrm{Pois}(20)$
```{r}
ppois(q = 12, lambda = 20)
```
Yes, it would be very unlikely to see so few deaths (0.03901) if nothing changed.

:::

## Pulmonary Disease, Environmental Health

Suppose the number of people seen for violent asthma attacks in the emergency ward of a hospital over a 1-day period is usually Poisson distributed with parameter λ = 1.5

::: {.panel-tabset} 
## 4.27 
What is the probability of observing 5 or more cases over a 2-day period?

## Solution
Let $X$ be the number observed over 2 days. Then $X \sim \mathrm{Pois}(3)$. We want $P(X \geq 5)$
```{r}
1 - ppois(q = 4, lambda = 3)
```

:::

On a particular 2-day period, the air-pollution levels increase dramatically and the distribution of attacks over a 1-day period is now estimated to be Poisson distributed with parameter λ = 3.

::: {.panel-tabset} 
## 4.28
Answer Problem 4.27 under these assumptions.

## Solution
Let $Y$ be the number observed over the 2 day period. Then $Y \sim \mathrm{Pois}(6)$. We want $P(Y \geq 5)$
```{r}
1 - ppois(q = 4, lambda = 6)
```

:::

::: {.panel-tabset} 
## 4.29 
If 10 days out of every year are high-pollution days, then what is the expected number of asthma cases seen in the emergency ward over a 1-year period? (Assume there are 365 days in a year.)

## Solution
10 days are $\mathrm{Pois}(3)$ and 355 days are $\mathrm{Pois}(1.5)$. So if $Z$is number in a year, we have $Z \sim \mathrm{Pois}(10 \times 3 + 355 \times 1.5) = \mathrm{Pois}(562.5)$. The expected number is 562.5.
:::

## Pulmonary Disease

Each year approximately 4% of current smokers attempt to quit smoking, and 50% of those who try to quit are successful in the sense that they abstain from smoking for at least 1 year from the date they quit.

::: {.panel-tabset} 
## 4.35 
What is the probability that a current smoker will quit for at least 1 year?

## Solution
$P(\mathrm{Quit}) = P(\mathrm{Quit}|\mathrm{Try})P(\mathrm{Try}) = 0.5 \times 0.04 = 0.02$.
:::


::: {.panel-tabset} 
## 4.36
What is the probability that among 100 current smokers, at least 5 will quit smoking for at least 1 year?

## Solution
Let $X$ be the number that quit. Then $X \sim \mathrm{Binom}(100, 0.02)$. We want $P(X \geq 5)$
```{r}
1 - pbinom(q = 4, size = 100, prob = 0.02)
```

:::

::: {.panel-tabset} 
## 4.37 
An educational program was conducted among smokers who attempt to quit to maximize the likelihood that such individuals would continue to abstain for the long term. Suppose that of 20 people who enter the program when they first stop smoking, 15 still abstain from smoking 1 year later. Can the program be considered successful?

## Solution
Assuming that it wasn't, $X \sim \mathrm{Binom}(20, 0.5)$. We use 0.5 instead of 0.02 because these are folks who have decided they want to quit. We want $P(X \geq 15)$
```{r}
1 - pbinom(q = 14, size = 20, prob = 0.5)
```

So we are very sure that the program is successful, since if it weren't we would only have a 2% probability of seeing as many or more successful quitters.
:::

## Infectious Disease
An outbreak of acute gastroenteritis occurred at a nursing home in Baltimore, Maryland, in December 1980. A total of 46 out of 98 residents of the nursing home became ill. People living in the nursing home shared rooms: 13 rooms contained 2 occupants, 4 rooms contained 3 occupants, and 15 rooms contained 4 occupants. One question that arises is whether or not a geographical clustering of disease occurred for persons living in the same room.

A summary of the number of affected people and the total number of people in a room is given in Table 4.3.

```{r}
#| echo: false
#| eval: true
#| message: false

library(gt)

# Create the data
df <- tribble(
  ~People_in_room, ~Total_rooms, ~`0_affected`, ~`1_affected`, ~`2_affected`, ~`3_affected`, ~`4_affected`,
  2, 13, 5, 4, 4, 0, 0,
  3, 4, 1, 2, 0, 1, 0,
  4, 15, 2, 4, 3, 5, 1
)

# Build gt table
df %>%
  gt() %>%
  tab_spanner(
    label = "Number of rooms with",
    columns = c(`0_affected`, `1_affected`, `2_affected`, `3_affected`, `4_affected`)
  ) %>%
  cols_label(
    People_in_room = "People in room",
    Total_rooms = "Total number of rooms",
    `0_affected` = "0 affected people",
    `1_affected` = "1 affected person",
    `2_affected` = "2 affected people",
    `3_affected` = "3 affected people",
    `4_affected` = "4 affected people"
  ) %>%
  tab_header(
    title = md("**Table 4.3**"),
    subtitle = "Number of affected people and total number of people in a room for an outbreak of acute gastroenteritis in a nursing home in Baltimore, Maryland"
  )

```


::: {.panel-tabset} 
## 4.47
If the binomial distribution holds, what is the probability distribution of the number of affected people in rooms with 2 occupants? That is, what is the probability of finding zero affected people? One affected person? Two affected people?

## Solution
Let's assume the probability of disease is the estimated probabiliy of disease 46/98 = 0.4694. Let $X$ be the number of affected out of 2. Then $X \sim \mathrm{Binom}(2, 0.4694)$. These probabilities are
```{r}
dbinom(x = 0:2, size = 2, prob = 0.4694)
```

:::

::: {.panel-tabset} 
## 4.48
Answer Problem 4.47 for the probability distribution of the number of affected people in rooms with 3 occupants.

## Solution
Let $Y$ be the number affected out of 3. Then $Y \sim \mathrm{Binom}(3, 0.4694)$. The probabilities are
```{r}
dbinom(x = 0:3, size = 3, prob = 0.4694)
```

:::

::: {.panel-tabset} 
## 4.49
Answer Problem 4.47 for the probability distribution of the number of affected people in rooms with 4 occupants.

## Solution
Let $Z$ be the number affected out of 4. Then $Z \sim \mathrm{Binom}(4, 0.4694)$. The probabilities are
```{r}
dbinom(x = 0:4, size = 4, prob = 0.4694)
```
:::

::: {.panel-tabset} 
## 4.50
One useful summary measure of geographical clustering is the number of rooms with 2 or more affected occupants. If the binomial distribution holds, what is the expected number of rooms with 2 or more affected occupants over the entire nursing home?

## Solution
There are 13 2-rooms, 4 3-rooms, and 15 4-rooms. The expected number of rooms with 2 or more affected folks is
$$
13 P(X \geq 2) + 4 P(Y \geq 2) + 15 P(Z \geq 2)
$$
where the random variables were defined in 4.47, 4.48, and 4.49. We calculate these probabilities as
```{r}
13 * (1 - pbinom(q = 1, size = 2, prob = 0.4694)) +
  4 * (1 - pbinom(q = 1, size = 3, prob = 0.4694)) +
  15 * (1 - pbinom(q = 1, size = 4, prob = 0.4694))
```

:::

## Cancer 2

The incidence rate of malignant melanoma in women ages 35–59 is approximately 11 new cases per 100,000 women per year. A study is planned to follow 10,000 women with excessive exposure to sunlight.

::: {.panel-tabset} 
## 4.52
What is the expected number of cases among 10,000 women over 4 years? (Assume no excess risk due to sunlight exposure.) 

## Solution
$$
10000 \text{ women} \times 4 \text{ years} \times \frac{11 \text{ cases}}{100000 \text{ women year}} = 4.4 \text{ cases}
$$
:::

::: {.panel-tabset} 
## 4.53
Suppose that 9 new cases are observed in this period. How unusual a finding is this?

## Solution
Let $X$ be the number of new cases over 4 years. The exact distribution is Binomial with size 10000 and probability 4 $\times$ 11 / 100000 = 0.00044. We want $P(X \geq 9)$
```{r}
1 - pbinom(q = 8, size = 10000, prob = 0.00044)
```
We can approximate this with a $\mathrm{Pois}(4.4)$
```{r}
1 - ppois(q = 8, lambda = 4.4)
```

So, somewhat unusual.
:::

## Cancer 3

A study was performed in Woburn, MA, looking at the rate of leukemia among children (≤ age 19) in the community in comparison to statewide leukemia rates. Suppose there are 12,000 children in the community that have lived there for a 10-year period and 12 leukemias have occurred in 10 years.

::: {.panel-tabset} 
## 4.54 
If the statewide incidence rate of leukemia in children is 5 events per 100,000 children per year (i.e., per 100,000 person-years) then how many leukemias would be expected in Woburn over the 10-year period if the statewide rates were applicable?

## Solution
$$
12000 \text{ children} \times 10 \text{ years} \times \frac{5 \text{ cases}}{100000 \text{ children year}} = 6
$$
:::

::: {.panel-tabset} 
## 4.55 
What is the probability of obtaining exactly 12 events over a 10-year period if statewide incidence rates were applicable?

## Solution
Let $X$ be the number of cases of 10 years. Then $X$ exactly follows a binomial with size 12000 and probability 10 $\times$ 5 / 100000 = 0.0005. We want $P(X = 12)$
```{r}
dbinom(x = 12, size = 12000, prob = 0.0005)
```
We can approximate this calculate with a $\mathrm{Pois}(6)$
```{r}
dpois(x = 12, lambda = 6)
```


:::

::: {.panel-tabset} 
## 4.56 
What is the probability of obtaining at least 12 events over the 10-year period if the statewide incidence rates were applicable?

## Solution
We want $P(X \geq 12)$. We can do this with binomial
```{r}
1 - pbinom(q = 11, size = 12000, prob = 0.0005)
```
or the Poisson
```{r}
1 - ppois(q = 11, lambda = 6)
```


:::

::: {.panel-tabset} 
## 4.57 
How do you interpret the results in Problem 4.56?

## Solution
It's rare to see so many. So there is likely extra risk in Woburn.
:::
