---
title: "Chapter 5 Practice Problems"
author: "David Gerard"
date: today
---

These practice problems mostly come from Rosner's publicly available study sheet at the books [companion website](https://www.cengage.com/cgi-wadsworth/course_products_wp.pl?fid=M20b&product_isbn_issn=9781305268920). The solutions are my own, since we differ slightly on what we are looking for in the solutions.

```{r}
#| echo: false
#| eval: true
#| message: false
library(tidyverse)
```

## Quantiles

::: {.panel-tabset} 
## 5.1
What are the deciles of the standard normal distribution; that is, the 10, 20, 30, ... , 90 percentiles?

## Solutions
```{r}
qnorm(p = seq(0.1, 0.9, by = 0.1))
```
:::

::: {.panel-tabset} 
## 5.2 
What are the quartiles of the standard normal distribution?

## Solutions
```{r}
qnorm(p = c(0.25, 0.5, 0.75))
```

:::

## Hypertension 1

Blood pressure (BP) in childhood tends to increase with age, but differently for boys and girls. Suppose that for both boys and girls, mean systolic blood pressure is 95 mm Hg at 3 years of age and increases 1.5 mm Hg per year up to the age of 13. Furthermore, starting at age 13, the mean increases by 2 mm Hg per year for boys and 1 mm Hg per year for girls up to the age of 18. Finally, assume that blood pressure is normally distributed and that the standard deviation is 12 mm Hg for all age-sex groups.

::: {.panel-tabset} 
## 5.3
What is the probability that an 11-year-old boy will have an SBP greater than 130 mm Hg?

## Solutions
Let $X$ be the SBP of a randomly selected 11 year old boy. Then $X$ is normal with mean 95 + 1.5 $\times$ 8 = 107 and standard deviation 12. We want $P(X > 130)$
```{r}
1 - pnorm(q = 130, mean = 107, sd = 12)
```

:::

::: {.panel-tabset} 
## 5.4
What is the probability that a 15-year-old girl will have an SBP between 100 and 120 mm Hg?

## Solutions
Let $Y$ be the SBP of a randomly selected 15-year old girl. Then $Y$ is normal with mean $95 + 1.5 \times 10 + 1 \times 2 = 112$ and standard deviation 12. We want $P(100 < Y < 120) = P(Y < 120) - P(Y < 100)$
```{r}
pnorm(q = 120, mean = 112, sd = 12) - pnorm(q = 100, mean = 112, sd = 12)
```

:::

::: {.panel-tabset} 
## 5.5
What proportion of 17-year-old boys have an SBP between 120 and 140 mm Hg?

## Solutions
Let $Z$ be the SBP of a randomly selected 17-year-old boy. Then $Z$ is normal with mean $95 + 1.5 \times 10 + 2 \times 4 = 118$ and standard deviation 12. We want $P(120 < Z < 140) = P(Z < 140) - P(Z < 120)$
```{r}
pnorm(q = 140, mean = 118, sd = 12) - pnorm(q = 120, mean = 118, sd = 12)
```

:::

::: {.panel-tabset} 
## 5.6
What is the probability that of 200 15-year-old boys, at least 10 will have an SBP of 130 mm Hg or greater?

## Hint
Let's first find the probability for a single 15-year-old boy.

## Solutions
Let's first find the probability for a single 15-year-old boy. Their SBP is normal with mean $95 + 1.5 \times 10 + 2 \times 2 = 114$ and sd 12. So the probability of their SBP being greater than 130 is
```{r}
1 - pnorm(q = 130, mean = 114, sd = 12)
```
Let $X$ now be the number of 15-year-old boys, out of 200, that have an SBP greater than 130. Then $X \sim \mathrm{Binom}(200, 0.09121)$. We want $P(X \geq 10)$
```{r}
1 - pbinom(q = 9, size = 200, prob = 0.09121)
```

:::

::: {.panel-tabset} 
## 5.7 
What level of SBP is at the 80th percentile for 7-year-old boys?

## Solutions
The SBP of a randomly sampled 7-year-old boy is normal with mean $95 + 1.5 \times 4 = 101$ and standard deviation 12. We want the 0.8 quantile of this distribution.
```{r}
qnorm(p = 0.8, mean = 101, sd = 12)
```

:::

::: {.panel-tabset} 
## 5.8 
What level of SBP is at the 70th percentile for 12-year-old girls?

## Solutions
The SBP of a randomly sampled 12-year-old girl is normal with mean $95 + 1.5 \times 9 = 108.5$ and standard deviation 12. We want the 0.7 quantile of this distribution.
```{r}
qnorm(p = 0.7, mean = 108.5, sd = 12)
```
:::

::: {.panel-tabset} 
## 5.9 
Suppose that a task force of pediatricians decides that children over the 95th percentile, but not over the 99th percentile, for their age-sex group should be encouraged to take preventive nonpharmacologic measures to reduce their blood pressure, whereas those children over the 99th percentile should receive antihypertensive drug therapy. Construct a table giving the appropriate BP levels to identify these groups for boys and girls for each year of age from 3 to 18.

## Hint
This is R heavy. You can skip for the exam. But it is great practice for a practical application of real statistics.

## Solutions
There are fancier ways to do this with less code. But I'll do it using just the techniques you know (except for `gt()`, which is a just function that makes a pretty table).
```{r}
#| message: false
library(tidyverse)
df_boy_young <- tibble(age = 3:13, sex = "M") |>
  mutate(mu = 95 + 1.5 * (age - 3), sd = 12)
df_boy_old <- tibble(age = 14:18, sex = "M") |>
  mutate(mu = 95 + 1.5 * 10 + 2 * (age - 13), sd = 12)
df_girl_young <- tibble(age = 3:13, sex = "F") |>
  mutate(mu = 95 + 1.5 * (age - 3), sd = 12)
df_girl_old <- tibble(age = 14:18, sex = "F") |>
  mutate(mu = 95 + 1.5 * 10 + 1 * (age - 13), sd = 12)
df <- bind_rows(df_boy_young, df_boy_old, df_girl_young, df_girl_old)
df |>
  mutate(
    bp95 = qnorm(p = 0.95, mean = mu, sd = sd),
    bp99 = qnorm(p = 0.99, mean = mu, sd = sd)
    ) |>
  select(Age = age, Sex = sex, bp95, bp99) |>
  gt::gt()
```
:::


## Cancer 1
The incidence of breast cancer in 40–49-year-old women is approximately 1 new case per 1000 women per year.

::: {.panel-tabset} 
## 5.10 
What is the incidence of breast cancer over 10 years in women initially 40 years old?

## Solutions
10 new cases per 1000 women per decade, which is the same as 1 case per 100 women, or 0.01.
:::

::: {.panel-tabset} 
## 5.11 
Suppose we are planning a study based on an enrollment of 10,000 women. What is the probability of obtaining at least 120 new breast-cancer cases over a 10-year follow-up period?

## Solutions
Let $X$ be the number of new breast-cancer cases. Then $X \sim \mathrm{Binom}(10000, 0.01)$. We want $P(X \geq 120)$, which is
```{r}
1 - pbinom(q = 119, size = 10000, prob = 0.01)
```
We can use a normal approximation too since $10000 \times 0.01 \times (1 - 0.01) = 99 \geq 5$. So $X$ is approximately normal with mean $np = 10000 \times 0.01 = 100$ and variance $10000 \times 0.01 \times (1 - 0.01) = 99$. Using a continuity correction, we calculate $P(X > 119.5)$
```{r}
1 - pnorm(q = 119.5, mean = 100, sd = sqrt(99))
```
These are pretty close.
:::

## Accident Epidemiology

Suppose the annual death rate from motor-vehicle accidents in 1980 was 10 per 100,000 in urban areas of the United States.

::: {.panel-tabset} 
## 5.12 
In an urban state with a population of 2 million, what is the probability of observing not more than 150 traffic fatalities in a given year?

## Solutions
The death rate is 
$$
2,000,000 \text{ folks} \times \frac{10 \text{ deaths}}{100,000 \text{ folks year}} = 200 \text{ deaths/year}
$$
Let $X$ be the number of deaths in a year. The exact distribution is $\mathrm{Binom}(2,000,000, 10/100,000)$ and we want $P(X \leq 150)$
```{r}
pbinom(q = 150, size = 2e6, prob = 1e-4)
```
The book wants you to use the normal approximation with mean 200 and standard deviation
```{r}
sqrt(2e6 * 1e-4 * (1-1e-4))
```
Using a continuity correction, we calculation $P(X \leq 150.5)$
```{r}
pnorm(q = 150.5, mean = 200, sd = 14.14)
```
It's actually very different (relatively speaking)! You should use the exact binomial probabilities in this case.
:::

In rural areas of the United States the annual death rate from motor-vehicle accidents is on the order of 100 per 100,000 population.

::: {.panel-tabset} 
## 5.13 
In a rural state with a population of 100,000, what is the probability of observing not more than 80 traffic fatalities in a given year?

## Solutions
Let $Y$ be the number of fatalities. The exact distribution is Binomial with size 100,000 and probability 100/100,000. We want $P(Y \leq 80)$
```{r}
pbinom(q = 80, size = 1e5, prob = 0.001)
```
Using a normal approximation with mean $np = 100,000 \times 0.001 = 100$ and variance $np(1-p) = 100,000 \times 0.001 \times (1 - 0.001) = 99.9$. Using a continuity correction, we calculation $P(Y < 80.5)$
```{r}
pnorm(80.5, mean = 100, sd = sqrt(99.9))
```


:::

::: {.panel-tabset} 
## 5.14 
What is the largest $x$ so that the probability of observing not more than $x$ traffic fatalities in a given year in a rural state with population 100,000 is less than 5%?

## Solutions
We want $x$ such that $P(X \leq x) < 0.05$. This is one less than the 0.05 quantile of the binomial distribution with size 100,000 and probability 0.001. "One less" because the 0.05 quantile is defined as the smallest value such that $P(X \leq x) \geq 0.05$
```{r}
qbinom(p = 0.05, size = 100000, prob = 0.001)
```
So, $x = 83$. Let's verify
```{r}
pbinom(q = 83, size = 100000, prob = 0.001)
pbinom(q = 84, size = 100000, prob = 0.001)
```
:::

::: {.panel-tabset} 
## 5.15
What is the smallest $x$ so that the probability of observing $x$ or more traffic fatalities in a given year, in an urban state with population 500,000 is less than 10%?

## Solutions
We want the 0.9 quantile of the binomial of size 500,000 and success probability 0.0001. This is exactly the definition of the quantile, the smallest $x$ such that $P(X \geq x) < 0.1$ is the same as the smallest $x$ such that $P(X \leq x) \geq 0.9$, which is the 0.9 quantile.
```{r}
qbinom(p = 0.9, size = 500000, prob = 0.0001)
```
So $x = 59$. Let's verify
```{r}
pbinom(q = 59, size = 500000, prob = 0.0001)
pbinom(q = 58, size = 500000, prob = 0.0001)
```

:::

## Obstetrics

Assume that birthweights are normally distributed with a mean of 3400 g and a standard deviation of 700 g.

::: {.panel-tabset} 
## 5.16 
Find the probability of a low-birthweight child, where low birthweight is defined as ≤ 2500 g.

## Solutions
We want $P(X \leq 2500)$ when $X \sim N(3400, 700^2)$
```{r}
pnorm(q = 2500, mean = 3400, sd = 700)
```

:::

::: {.panel-tabset} 
## 5.17 
Find the probability of a very low birthweight child, where very low birthweight is defined as ≤ 2000 g.

## Solutions
Similar to 5.16
```{r}
pnorm(q = 2000, mean = 3400, sd = 700)
```

:::

::: {.panel-tabset} 
## 5.18 
Assuming that successive deliveries by the same woman have the same probability of being low birthweight, what is the probability that a woman with exactly 3 deliveries will have 2 or more low-birthweight deliveries?

## Solutions
Let $X$ be the number of low-birthweights out of 3. Then $X \sim \mathrm{Binom}(3, 0.09927)$, where the success probability was calculated in 5.16. We want $P(X \geq 2)$
```{r}
1 - pbinom(q = 1, size = 3, prob = 0.09927)
```
:::

## Renal Disease

The presence of bacteria in a urine sample (bacteriuria) is sometimes associated with symptoms of kidney disease. Assume that a determination of bacteriuria has been made over a large population at one point in time and that 5% of those sampled are positive for bacteriuria.

::: {.panel-tabset} 
## 5.19 
Suppose that 500 people from this population are sampled. What is the probability that 30 or more people would be positive for bacteriuria?

## Solutions
Let $X$ be the number out of 500 that are positive. Then $X \sim \mathrm{Binom}(500, 0.05)$. We want $P(X \geq 30)$
```{r}
1 - pbinom(q = 30, size = 500, prob = 0.05)
```
We can use the normal approximation here since $500 \times 0.05 \times (1 - 0.05) = 23.75 > 5$. Then $X$ is about normal with mean $np = 500 \times 0.05 = 25$ and variance 23.75. Using a continutity correction, we calculation $P(X > 29.5)$
```{r}
1 - pnorm(q = 29.5, mean = 25, sd = sqrt(23.75))
```
:::

## Ophthalmology 1

A study was conducted among patients with retinitis pigmentosa, an ocular condition where pigment appears over the retina, resulting in substantial loss of vision in many cases. The study was based on 94 patients who were seen annually at a baseline visit and at three annual follow-up visits. In this study, 90 patients provided visual-field measurements at each of the four examinations and are the subjects of the following data analyses. Visual field was transformed to the log scale to better approximate normality and yielded the data given in Table 5.1.

```{r}
#| eval: true
#| echo: false
#| message: false
library(gt)
# Data
df <- tribble(
  ~`Year of examination`, ~Mean, ~`Standard deviation`, ~n,
  "Year 0 (baseline)", 8.15, 1.23, 90,
  "Year 3", 8.01, 1.33, 90,
  "Year 0–year 3", 0.14, 0.66, 90
)

# Create gt table
df %>%
  gt() %>%
  tab_header(
    title = md("**Table 5.1** Visual-field measurements in retinitis-pigmentosa patients")
  ) %>%
  fmt_number(
    columns = c(Mean, `Standard deviation`),
    decimals = 2
  ) %>%
  cols_label(
    `Year of examination` = "Year of examination",
    Mean = html("Mean"),
    `Standard deviation` = "Standard deviation",
    n = html("<i>n</i>")
  )
```

::: {.panel-tabset} 
## 5.20
Assuming that change in visual field over 3 years is normally distributed when using the log scale, what is the proportion of patients who showed a decline in visual field over 3 years?

## Solutions
Let $X$ be the visual acuity change of a randomly selected patient. Then $X$ is normal with mean 0.14 and standard deviation 0.66. We want $P(X > 0)$ (greater than because it's year 0 minus year 3, so the decline happens if the $X$ is positive).
```{r}
1 - pnorm(q = 0, mean = 0.14, sd = 0.66)
```

:::

::: {.panel-tabset} 
## 5.21 
What percentage of patients would be expected to show a decline of at least 20% in visual field over 3 years? (Note: In the ln scale this is equivalent to a decline of at least log(1/0.8) = 0.223).

## Solutions
Using the notation from 5.22, we want $P(X > 0.223)$
```{r}
1 - pnorm(q = 0.223, mean = 0.14, sd = 0.66)
```
:::

::: {.panel-tabset} 
## 5.22 
Answer Problem 5.21 for a 50% decline over 3 years.

## Solutions
The lower bound is
```{r}
log(1 / 0.5)
```
So we want $P(X > 0.6931)$
```{r}
1 - pnorm(q = 0.6931, mean = 0.14, sd = 0.66)
```
:::

## Cancer 2

A study of the Massachusetts Department of Health found 46 deaths due to cancer among women in the city of Bedford, MA over the period 1974–1978, where 30 deaths had been expected from statewide rates.

::: {.panel-tabset} 
## 5.23 
Write an expression for the probability of observing exactly $k$ deaths due to cancer over this period if the statewide rates are correct.

## Solutions
Let $X$ be the number of deaths. Then $X$ is Poisson with mean $30$. You can write out the PMF here
$$
\frac{30^ke^{-30}}{k!}.
$$
:::

::: {.panel-tabset} 
## 5.24 
Can the occurrence of 46 deaths due to cancer be attributed to chance? Specifically, what is the probability of observing at least 46 deaths due to cancer if the statewide rates are correct?

## Solutions
Using the same notation as in 5.23, we want $P(X \geq 46)$
```{r}
1 - ppois(q = 45, lambda = 30)
```
This is very rare, and so we strong evidence that the rates are different than the specified state-wide rates.

We can also use the normal approximation of the Poisson with mean and variance of 30. Using a contintuity correction, we want $P(X > 45.5)$
```{r}
1 - pnorm(q = 45.5, mean = 30, sd = sqrt(30))
```
:::

## Nutrition

The distribution of serum levels of alpha tocopherol (serum vitamin E) is approximately normal with mean 860 µg dL and standard deviation 340 µg dL.

::: {.panel-tabset} 
## 5.27
What percentage of people have serum alpha tocopherol levels between 400 and 1000 µg dL ?

## Solutions
Let $X$ be normal with mean 860 and standard deviation 340. We want $P(400 < X < 1000)$
```{r}
pnorm(1000, mean = 860, sd = 340) - pnorm(q = 400, mean = 860, sd = 340)
```

:::

::: {.panel-tabset} 
## 5.28
Suppose a person is identified as having toxic levels of alpha tocopherol if his or her serum level is > 2000 µg dL. What percentage of people will be so identified?

## Solutions
Using the same notation as 5.27, we want $P(X > 2000)$
```{r}
1 - pnorm(2000, mean = 860, sd = 340)
```

:::

::: {.panel-tabset} 
## 5.29
A study is undertaken for evidence of toxicity among 2000 people who regularly take vitamin-E supplements. The investigators found that 4 people have serum alpha tocopherol levels > 2000 µg dL. Is this an unusual number of people with toxic levels of serum alpha tocopherol?

## Solutions
Let $Y$ be the number out of 2000 with levels > 2000 µg. Then $Y$ is Binomial with size 2000 and probability 0.0003998 (from 5.28). We want $P(Y \geq 4)$
```{r}
1 - pbinom(q = 3, size = 2000, prob = 0.0003998)
```
This is very unusual.
:::


## Pulmonary Disease

::: {.panel-tabset} 
## 5.32 
The usual annual death rate from asthma in England over the period 1862–1962 for people aged 5–34 was approximately 1 per 100,000. Suppose that in 1963 twenty deaths were observed in a group of 1 million people in this age group living in Britain. Is this number of deaths inconsistent with the preceding 100-year rate? In particular, what is the probability of observing 20 or more deaths in 1 year in a group of 1 million people? Note: This finding is both statistically and medically interesting, since it was found that the excess risk could be attributed to certain aerosols used by asthmatics in Britain during the period 1963–1967. The rate returned to normal during the period 1968–1972, when these types of aerosols were no longer used.

## Solutions
Let $X$ be the number of deaths in a year. Then $X$ is poisson with mean 1,000,000 $\times$ 1/100,000 = 10. We want $P(X \geq 20)$.
```{r}
1 - ppois(q = 19, lambda = 10)
```
This would be very rare.

Note that the exact distribution is Binomial with size 1,000,000 and probability 1/100,000. But we get the same numbers:
```{r}
1 - pbinom(q = 19, size = 1e6, prob = 1e-5)
```
:::

## Hypertension 2

Blood-pressure measurements are known to be variable, and repeated measurements are essential to accurately characterize a person’s BP status. Suppose a person is measured on $n$ visits with $k$ measurements per visit and the average of all $nk$ diastolic blood pressure (DBP) measurements ($\bar{x}$) is used to classify a person as to BP status. Specifically, if $\bar{x}$ ≥ 95 mm Hg, then the person is classified as hypertensive; if $\bar{x}$ < 90 mm Hg, then the person is classified as normotensive; and if $\bar{x}$ ≥ 90 mm Hg, and < 95 mm Hg, the person is classified as borderline. It is also assumed that a person's "true" blood pressure is $\mu$ , representing an average over a large number of visits with a large number of measurements per visit, and that $\bar{X}$ is normally distributed with mean $\mu$ and variance = $27.7/n + 7.9/(nk)$

::: {.panel-tabset} 
## 5.33 
If a person’s true diastolic blood pressure is 100 mm Hg, then what is the probability that the person will be classified accurately (as hypertensive) if a single measurement is taken at 1 visit?

## Solutions
We are given that $\mu = 100$ and $n = 1$ and $k = 1$, so the variance is $27.7/1 + 7.9/(1\times 1) = 35.6$. Let $X$ be the person's single measurement, then we want $P(X \geq 95)$
```{r}
1 - pnorm(q = 95, mean = 100, sd = sqrt(35.6))
```
So there is only a 79.9% chance of correctly classifying that person on one visit.
:::

::: {.panel-tabset} 
## 5.34 
Is the probability in Problem 5.33 a measure of sensitivity, specificity, or predictive value?

## Solutions
Sensitifity. This is the probability of the test being positive given the disease is positive.
:::

::: {.panel-tabset} 
## 5.35 
If a person’s true blood pressure is 85 mm Hg, then what is the probability that the person will be accurately classified (as normotensive) if 3 measurements are taken at each of 2 visits?

## Solutions
We are given $\mu = 85$ and $k = 3$ and $n = 2$. So the variance is $27.7/2 + 7.9/(2\times 3) = 15.17$. So the probability of being accurately classified at is $P(X \leq 90)$
```{r}
pnorm(q = 90, mean = 85, sd = sqrt(15.17))
```
So there is a 90% chance of being correctly classified.
:::

::: {.panel-tabset} 
## 5.36 
Is the probability in Problem 5.35 a measure of sensitivity, specificity, or predictive value?

## Solutions
Specificity. This is the probability of the test being negative given the disease is negative.
:::

::: {.panel-tabset} 
## 5.37 
Suppose we decide to take 2 measurements per visit. How many visits are needed so that the sensitivity and specificity in Problems 5.33 and 5.35 would each be at least 95%?

## Solutions
The book's solution is a lot of math, which is great, and something we'll go through later. But we can do a quick and dirty version on R by just trying out some values of n:
```{r}
k <- 2
n <- 3
sigma <- sqrt(27.7 / n + 7.9/(n *  k))
1 - pnorm(q = 95, mean = 100, sd = sigma)
pnorm(q = 90, mean = 85, sd = sigma)

n <- 4
sigma <- sqrt(27.7 / n + 7.9/(n *  k))
1 - pnorm(q = 95, mean = 100, sd = sigma)
pnorm(q = 90, mean = 85, sd = sigma)
```
So, 4 visits.
:::


## Hypertension 3

A study is planned to look at the effect of sodium restriction on lowering blood pressure. Nutritional counseling sessions are planned for the participants to encourage dietary sodium restriction. An important component in the study is validating the extent to which individuals comply with a sodium-restricted diet. This is usually accomplished by obtaining urine specimens and measuring sodium in the urine.

Assume that in free-living individuals (individuals with no sodium restriction) 24-hour urinary sodium excretion is normally distributed with mean 160.5 mEq/24 hrs and standard deviation = 57.4 mEq/24 hrs.


::: {.panel-tabset} 
## 5.38 
If 100 mEq/24 hr is the cutoff value chosen to measure compliance, then what percentage of noncompliant individuals—that is, individuals who do not restrict sodium—will have a 24-hour sodium level below this cutoff point?

## Solutions
Let $X$ be normal with mean 160.5 and standard deviation 57.4. Then we want $P(X < 100)$
```{r}
pnorm(q = 100, mean = 160.5, sd = 57.4)
```

:::

Suppose that in an experimental study it is found that people who are on a sodium-restricted diet have 24-hour sodium excretion that is normally distributed with mean 57.5 mEq/24 hrs and standard deviation of 11.3 mEq/24 hrs.

::: {.panel-tabset} 
## 5.39
What is the probability that a person who is on a sodium-restricted diet will have a 24-hour urinary sodium level above the cutoff point (100 mEq/24 hrs)? 

## Solutions
Let $X$ be normal with mean 57.5 and standard deviation 11.3. Then we want $P(X > 100)$.
```{r}
1 - pnorm(q = 100, mean = 57.5, sd = 11.3)
```

:::

::: {.panel-tabset} 
## 5.40 
Suppose the investigators in the study wish to change the cutoff value (from 100 mEq/24 hrs) to another value such that the misclassification probabilities in Problems 5.38 and 5.39 are the same. What should the new cutoff value be, and what are the misclassification probabilities corresponding to your answers to Problems 5.38 and 5.39?

## Solutions

Let $x$ be the cutoff value. Then we can re-write 5.38 at
$$
\Phi\left(\frac{x - 160.5}{57.4} \right),
$$
where $\Phi(\cdot)$ is the CDF of the standard normal. We can also re-write 5.39 as
$$
1 - \Phi\left(\frac{x - 57.5}{11.3} \right) = \Phi\left(-\frac{x - 57.5}{11.3} \right)
$$
So we want to solve for $x$ in the equation
$$
\Phi\left(\frac{x - 160.5}{57.4} \right) = \Phi\left(-\frac{x - 57.5}{11.3}\right)
\Leftrightarrow \frac{x - 160.5}{57.4} = -\frac{x - 57.5}{11.3}
$$
This can be solved using basic algebra. You end up getting $x = 74.4$.

You won't get such lengthy algebra on an exam. But you should absolutely know that
`pnorm(x, μ, σ)` is the exact same thing as $\Phi\left(\frac{x - \mu}{\sigma}\right)$
:::

## Alcohol and Hypertension.

Table 5.2 shows data reported relating alcohol consumption and blood-pressure level among 18–39-year-old males. The "Elevated blood pressure" column means either SBP ≥ 140 mm Hg or DBP ≥ 90 mm Hg (or both).

```{r}
#| echo: false
#| eval: true
#| message: false

library(gt)

# Create the data frame
df <- tibble::tibble(
  Group = c("Nondrinker", "Heavy drinker"),
  SBP_Mean = c(120.2, 123.5),
  SBP_SD   = c(10.7, 12.8),
  SBP_n    = c(96, 124),
  DBP_Mean = c(74.8, 78.5),
  DBP_SD   = c(10.2, 9.2),
  DBP_n    = c(96, 124),
  Elevated_BP = c("11.5%", "17.7%")
)

# Build the gt table
df %>%
  gt(rowname_col = "Group") %>%
  tab_header(
    title = md("**TABLE 5.2**  The relationship of blood pressure level to drinking behavior")
  ) %>%
  tab_spanner(
    label = "Systolic blood pressure (SBP)",
    columns = c(SBP_Mean, SBP_SD, SBP_n)
  ) %>%
  tab_spanner(
    label = "Diastolic blood pressure (DBP)",
    columns = c(DBP_Mean, DBP_SD, DBP_n)
  ) %>%
  tab_spanner(
    label = "Elevated blood pressure",
    columns = Elevated_BP
  ) %>%
  cols_label(
    SBP_Mean = "Mean",
    SBP_SD   = "sd",
    SBP_n    = "n",
    DBP_Mean = "Mean",
    DBP_SD   = "sd",
    DBP_n    = "n",
    Elevated_BP = html("&nbsp;")
  ) %>%
  tab_options(
    table.font.size = "small",
    data_row.padding = px(4)
  )
```


::: {.panel-tabset} 
## 5.42 
If we assume that the distributions of SBP are normal, then what proportion of nondrinkers have SBP ≥ 140 mm Hg?

## Solutions
Let $X$ be a randomly sampled non-drinker. then $X$ is normal with mean 120.2 and standard deviation 10.7. We want $P(X \geq 140)$
```{r}
1 - pnorm(q = 140, mean = 120.2, sd = 10.7)
```

:::

::: {.panel-tabset} 
## 5.43
What proportion of heavy drinkers have SBP ≥ 140 mm Hg?

## Solutions
Let $Y$ be a randomly selected heavy drinker. Then $X$ is normal with mean 123.5 and standard deviation 12.8. We want $P(Y \geq 140)$.
```{r}
1 - pnorm(q = 140, mean = 123.5, sd = 12.8)
```

:::

::: {.panel-tabset} 
## 5.44
What proportion of heavy drinkers have “isolated systolic hypertension”; i.e., SBP ≥ 140 mm Hg but DBP < 90 mm Hg?

## Hint
$$
P(A\cap \bar{B}) = P(A \cup B) - P(B)
$$
Draw it out to convince yourself.

## Solutions
Let $A$ be SBP > 140 and $B$ be DBP > 90. We want $P(A \cap \bar{B})$. We are given that $P(A \cup B) = 0.177$. We can calculate $P(B)$ from a normal with mean 78.5 and standard deviation 9.2 as
```{r}
1 - pnorm(q = 90, mean = 78.5, sd = 9.2)
```
Using the hint, we have
$$
P(A\cap \bar{B}) = P(A \cup B) - P(B) = 0.177 - 0.1056 = 0.0714
$$
:::

::: {.panel-tabset} 
## 5.45 
What proportion of heavy drinkers have "isolated diastolic hypertension"; i.e., DBP ≥ 90 mm Hg but SBP < 140 mm Hg?

## Solutions
Using the same notation as 5.44, we want $P(\bar{A} \cap B)$. $P(A) = 0.09869$ from 5.43. We are also given  $P(A \cup B) = 0.177$. Using the same strategy as 5.44, we have
$$
P(\bar{A} \cap B) = P(A \cup B) - P(A) = 0.177 - 0.09869 = 0.07831
$$
:::

## Hypertension 4

The Pediatric Task Force Report on Blood Pressure Control in Children reports blood-pressure norms for children by age and sex group. The mean ± standard deviation for 17-year-old boys for diastolic blood pressure (DBP) is 63.7 ± 11.4 mm Hg based on a large sample.

::: {.panel-tabset} 
## 5.46
Suppose the 90th percentile of the distribution is the cutoff for elevated BP. If we assume that the distribution of blood pressure for 17-year-old boys is normally distributed, then what is the cutoff in mm Hg?

## Solutions
This is the 0.9 quantile of a normal distribution with mean 63.7 and standard deviation 11.4.
```{r}
qnorm(p = 0.9, mean = 63.7, sd = 11.4)
```

:::

::: {.panel-tabset} 
## 5.47 
Another approach for defining elevated BP is to use 90 mm Hg as the cutoff (the standard for elevated adult DBP). What percentage of 17-year-old boys would have elevated BP using this approach?

## Solutions
We want the $P(X \geq 90)$ where $X$ is normal with mean 63.7 and standard deviation 11.4.
```{r}
1 - pnorm(q = 90, mean = 63.7, sd = 11.4)
```
Only about 1% of 17-year-old boys.
:::

::: {.panel-tabset} 
## 5.48 
Suppose there are 200 17-year-old boys in the 11th grade of whom 25 have elevated BP using the criteria in Problem 5.46. Are there an unusually large number of boys with elevated BP? Why or why not?

## Solutions
Let $X$ be the number of boys with elevated blood pressure out of 200. Then we would expect $X \sim \mathrm{Binom}(200, 0.1)$ (only 10% have high blood pressure on average). We want $P(X \geq 25)$.
```{r}
1 - pbinom(q = 24, size = 200, prob = 0.1)
```
Not that unusual.

:::

## Environmental Health

::: {.panel-tabset} 
## 5.49 
A study was conducted relating particulate air pollution and daily mortality in Steubenville, Ohio. On average over the last 10 years, there have been 3 deaths per day. Suppose that on 90 high-pollution days (where the total suspended particulates are in the highest quartile among all days) the death rate is 3.2 deaths per day or 288 deaths observed over the 90 high-pollution days. Are there an unusual number of deaths on high-pollution days?

## Solutions
Let $X$ be the number of deaths in 90 days. Then we expect $X \sim \mathrm{Pois}(90 \times 3) = \mathrm{Pois}(270)$. We want $P(X \geq 288)$.
```{r}
1 - ppois(q = 287, lambda = 270)
```
Not that unusual.

We can also use a normal approximate with mean 270 and variance 270. Using a continuity correction, we calculate $P(X \geq 287.5)$
```{r}
1 - pnorm(q = 287.5, mean = 270, sd = sqrt(270))
```
:::


## Ophthalmology 2

In ophthalmology, it is customary to obtain measurements on both the right eye (OD) and the left eye (OS). Suppose intraocular pressure (IOP) is measured on each eye and an average of the OD and OS values (denoted by the OU value) is computed. We wish to compute the variance of the IOP OU values for an individual person. Assume that the standard deviation is 4 mm Hg for the OD values and 4 mm Hg for the OS values (as estimated from the sample standard deviation of IOP from right and left eyes, respectively, from a sample of 100 people).

::: {.panel-tabset} 
## 5.53 
If the OD and OS values are assumed to be independent, then what is the variance of the IOP OU value for individual persons?

## Solutions
Let $X$ be OD value with $\mathrm{var}(X) = 4^2 = 16$ and $Y$ be the OS value wiht $\mathrm{var}(Y) = 4^2 = 16$. We have the OU value as $Z = (X + Y)/2)$. We want
$$
\mathrm{Z} = \frac{1}{4}\mathrm{var}(X) + \frac{1}{4}\mathrm{var}(Y) = 16/4 + 16/4 = 8.
$$
:::
