---
title: "Chapter 7 Practice Problems"
author: "David Gerard"
date: today
---

These practice problems mostly come from Rosner's publicly available study sheet at the books [companion website](https://www.cengage.com/cgi-wadsworth/course_products_wp.pl?fid=M20b&product_isbn_issn=9781305268920). The solutions are my own, since we differ slightly on what we are looking for in the solutions.

```{r setup}
#| echo: false
#| eval: true
#| message: false
library(tidyverse)
library(gt)
library(broom)
plt_t <- function(
    mu = 0, 
    sig = 1, 
    lb = -Inf,
    ub = Inf,
    df = Inf, 
    rng = c(-3, 3), 
    two_sided = FALSE,
    col = "#E69F00",
    lwd = 1) {
  tibble(x = seq(mu + rng[[1]] * sig, mu + rng[[2]] * sig, length.out = 500)) |>
    mutate(y = dt(x = (x - mu) / sig, df = df)) ->
    df1
  df1 |>
    filter(x > lb, x < ub) ->
    df2
  ggplot() +
    geom_line(data = df1, mapping = aes(x = x, y = y), linewidth = lwd) +
    geom_area(data = df2, mapping = aes(x = x, y = y), fill = col) +
    geom_line(data = df2, mapping = aes(x = x, y = y), color = col, linewidth = lwd) +
    theme_classic() +
    theme(axis.title = element_blank()) ->
    pl
  
  if (two_sided) {
    df1 |>
      filter(x > -ub, x < -lb) ->
      df3
    pl <- pl + 
      geom_area(data = df3, mapping = aes(x = x, y = y), fill = col) +
      geom_line(data = df3, mapping = aes(x = x, y = y), color = col, linewidth = lwd)
  }
  pl
}
plt_binom <- function(
    size, 
    prob = 0.5, 
    lb = -Inf,
    ub = Inf, 
    rng = c(0, size),
    col = "#E69F00",
    valmax = NULL,
    lwd = 1) {
  
  tibble(x = seq(rng[[1]], rng[[2]], by = 1)) |>
    mutate(
      pmf = stats::dbinom(x = x, size = size, prob = prob)
      ) ->
    df
  
  if (is.null(valmax)) {
    df |>
      mutate(col = (x >= lb & x <= ub)) ->
      df
  } else {
    df |>
      mutate(col = pmf <= stats::dbinom(x = valmax, size = size, prob = prob)) ->
      df
  }
  df |>
    ggplot(aes(x = x, xend = x, yend = pmf, color = col)) +
    geom_segment(y = 0, lwd = lwd) +
    theme_classic() +
    theme(axis.title = element_blank()) +
    scale_color_manual(values = c("black", col), guide = "none") +
    scale_x_continuous(breaks = seq(rng[[1]], rng[[2]]))
}
plt_pois <- function(
    mu, 
    lb = -Inf,
    ub = Inf, 
    rng = c(0, 2 * mu),
    col = "#E69F00",
    valmax = NULL,
    lwd = 1) {
  
  tibble(x = seq(rng[[1]], rng[[2]], by = 1)) |>
    mutate(
      pmf = stats::dpois(x = x, lambda = mu)
      ) ->
    df
  
  if (is.null(valmax)) {
    df |>
      mutate(col = (x >= lb & x <= ub)) ->
      df
  } else {
    df |>
      mutate(col = pmf <= stats::dpois(x = valmax, lambda = mu)) ->
      df
  }
  df |>
    ggplot(aes(x = x, xend = x, yend = pmf, color = col)) +
    geom_segment(y = 0, lwd = lwd) +
    theme_classic() +
    theme(axis.title = element_blank()) +
    scale_color_manual(values = c("black", col), guide = "none") +
    scale_x_continuous(breaks = seq(rng[[1]], rng[[2]]))
}
```

## Nutrition 1

As part of a dietary-instruction program, ten 25–34-year-old males adopted a vegetarian diet for 1 month. While on the diet, the average daily intake of linoleic acid was 13 g with standard deviation = 4 g.

::: {.panel-tabset}
## 7.1 
If the average daily intake among 25–34-year-old males in the general population is 15 g, then, using a significance level of .05, test the hypothesis that the intake of linoleic acid in this group is lower than that in the general population.

## Solution
Let $X_i$ be the average daily intake of linoleic acid of individual $i$. We assume that $X_i \sim N(\mu, \sigma^2)$. We are testing $H_0: \mu = 15$ versus $H_A: \mu < 15$. We are told that $\bar{x} = 13$, $s = 4$, and $n = 10$. We first calculate the $t$-statistic
$$
t = \frac{\bar{x} - \mu_0}{s/\sqrt{n}} = \frac{13 - 15}{4 / \sqrt{10}}
$$
Numerically
```{r}
(13 - 15) / (4 / sqrt(10))
```
For the critical value method, we compare this to $t_{n-1, \alpha}$ where $\alpha = 0.05$
```{r}
qt(p = 0.05, df = 10 - 1)
```
Recall that since we are testing if the mean is lower than some value, evidence against the null is for very negative $t$-statistics. Since $t > t_{n-1, \alpha}$ (-1.581 > -1.833), we fail to reject $H_0$, and conclude that we do not have evidence that the average linoleic acid is lower than the general population.
:::

::: {.panel-tabset}
## 7.2 
Compute a p-value for the hypothesis test in Problem 7.1.

## Solution
We could also calculate the $p$-value, $P(T < -1.581)$ where $T \sim t_{n-1}$
```{r}
#| echo: false
plt_t(ub = -1.581, df = 9) +
  geom_vline(xintercept = -1.581, lty = 2)
```
Numerically
```{r}
pt(q = -1.581, df = 9)
```
Since the $p$-value is greater than 0.05, we again fail to reject at level 0.05 and conclude that we do not have evidence that the average linoleic acid is lower than the general population.
:::

As part of the same program, eight 25–34-year-old females report an average daily intake of saturated fat of 11 g with standard deviation = 11 g while on a vegetarian diet.

::: {.panel-tabset}
## 7.3 
If the average daily intake of saturated fat among 25–34-year-old females in the general population is 24 g, then, using a significance level of .01, test the hypothesis that the intake of saturated fat in this group is lower than that in the general population.

## Solution
Let $X_i$ be the average daily intake of saturated fat of individual $i$. We assume that $X_i \sim N(\mu, \sigma^2)$. We are testing $H_0: \mu = 24$ versus $H_A: \mu < 24$. We are told that $\bar{x} = 11$, $s = 11$, and $n = 8$. We first calculate the $t$-statistic
$$
t = \frac{\bar{x} - \mu_0}{s/\sqrt{n}} = \frac{11 - 24}{11 / \sqrt{8}}
$$
Numerically
```{r}
(11 - 24) / (11 / sqrt(8))
```
For the critical value method, we compare this to $t_{n-1, \alpha}$ where $\alpha = 0.01$
```{r}
qt(p = 0.01, df = 8 - 1)
```
Recall that since we are testing if the mean is lower than some value, evidence against the null is for very negative $t$-statistics. Since $t < t_{n-1, \alpha}$ (-3.343 < -2.998), we reject $H_0$, and conclude that we have evidence that the average saturated fat intake is lower in this group.
:::

::: {.panel-tabset}
## 7.4 
Compute a p-value for the hypothesis test in Problem 7.3.

## Solution
We could also calculate the $p$-value, $P(T < -3.343)$ where $T \sim t_{n-1}$
```{r}
#| echo: false
plt_t(ub = -3.343, rng = c(-4, 4), df = 7) +
  geom_vline(xintercept = -3.343, lty = 2)
```
Numerically
```{r}
pt(q = -3.343, df = 7)
```
Since the $p$-value is less than 0.01, we reject the null and conclude that the average saturated fat intake for this group is lower than the general population.
:::

::: {.panel-tabset}
## 7.5 
What is the relationship between your answers to Problems 7.3 and 7.4?

## Solution
- Let $t$ be our observed $t$-statistic. 
- Let $t_{n-1, \alpha}$ be the $\alpha$ quantile of a $t$-distribution with $n-1$ degrees of freedom.
- Let $T$ be a random variable that follows a $t$-distribution with $n-1$ degrees of freedom. That is, $T \sim t_{n-1}$.

Then $t$ is less than $t_{n-1, \alpha}$ (critical value method) if and only if $P(T < t) < \alpha$ ($p$-value method).
:::

Suppose we are uncertain what effect a vegetarian diet will have on the level of linoleic-acid intake in Problem 7.1.

::: {.panel-tabset}
## 7.6
What are the null and alternative hypotheses in this case?

## Solution
Using the same notation as in 7.1, we are testing

- $H_0: \mu = 15$
- $H_A: \mu \neq 15$
:::

::: {.panel-tabset}
## 7.7
Compare the mean level of linoleic acid in the vegetarian population with that of the general population under the hypotheses in Problem 7.6. Report a $p$-value.

## Solution
We use the same $t$-statistic as before
$$
t = \frac{\bar{x} - \mu_0}{s/\sqrt{n}} = \frac{13 - 15}{4 / \sqrt{10}} = -1.581
$$
But now "more extreme" means just further away from 0. For the critical value method, we compare our observed $t$-statistic to $t_{n-1, 1 - \alpha/2}$
```{r}
qt(p = 1 - 0.05 / 2, df = 10 - 1)
```
Since $|t| < t_{n-1,1 - \alpha/2}$ (1.581 < 2.262), we again fail to reject $H_0$ at significance level 0.05. More commonly, we would calculate a $p$-value as $P(|T| > 1.581)$
```{r}
#| echo: false
plt_t(ub = -1.581, df = 9, two_sided = TRUE) +
  geom_vline(xintercept = -1.581, lty = 2)
```
In R
```{r}
2 * pt(q = -1.581, df = 10 - 1)
```
Since the $p$-value is 0.1483, we fail to reject $H_0$ at significance level 0.05 and conclude that we do not have evidence that mean linoleic acid is different in this group than in the general population.
:::

::: {.panel-tabset}
## 7.10
Suppose we are uncertain what effect a vegetarian diet will have on saturated fat intake. Compute a two-sided p-value for the hypothesis testing situation mentioned in Problem 7.3.

## Solution
We use the same $t$-statistic as before
$$
t = \frac{\bar{x} - \mu_0}{s/\sqrt{n}} = \frac{11 - 24}{11 / \sqrt{8}} = -3.343
$$
But now "more extreme" means just further away from 0. For the critical value method, we compare our observed $t$-statistic to $t_{n-1, 1 - \alpha/2}$
```{r}
qt(p = 1 - 0.01 / 2, df = 8 - 1)
```
Since $|t| < t_{n-1,1 - \alpha/2}$ (3.343 < 3.499), this time we fail to reject $H_0$ at significance level 0.01. More commonly, we would calculate a $p$-value as $P(|T| > 3.343)$
```{r}
#| echo: false
plt_t(ub = -3.343, df = 7, two_sided = TRUE, rng = c(-4, 4)) +
  geom_vline(xintercept = -3.343, lty = 2)
```
In R
```{r}
2 * pt(q = -3.343, df = 8 - 1)
```
Since the $p$-value is 0.01237 (larger than 0.01), we fail to reject $H_0$ at significance level 0.01 and conclude that we do not have evidence that mean saturated fat is different in this group than in the general population.
:::

::: {.panel-tabset}
## 7.11
Compute a lower one-sided 95% CI for the true mean intake of linoleic acid in the vegetarian population depicted in Problem 7.1.

## Solution
We need an upper bound of the form
$$
\bar{x} + t_{n-1, 1-\alpha}\frac{s}{\sqrt{n}}
$$
We were given all terms except $t_{n-1,1-\alpha}$, where $\alpha = 0.05$
```{r}
qt(p = 1 - 0.05, df = 10 - 1)
```
Plugging in values, we get
$$
13 + 1.833 \times 4 / \sqrt{10}
$$
Numerically
```{r}
13 + 1.833 * 4 / sqrt(10)
```
So the one-sided 95% CI is
$$
\mu \leq 15.32
$$
:::

::: {.panel-tabset}
## 7.12
How does your answer to Problem 7.11 relate to your answer to Problem 7.1?

## Solution
Since the 95% CI include 15 (from 7.11), we would fail to reject at level 0.05 for $H_0: \mu = 15$ versus $H_A: \mu < 15$ (from 7.1).
:::

## Occupational Health 1

Suppose that 28 cancer deaths are noted among workers exposed to asbestos in a building-materials plant from 1981–1985. Only 20.5 cancer deaths are expected from statewide cancer-mortality rates.

::: {.panel-tabset}
## 7.17
What is the estimated SMR for total cancer mortality?

## Solution
SMR is
$$
\frac{\text{observed}}{\text{expected}}\times 100\% = \frac{28}{20.5}\times 100\%
$$
```{r}
28 / 20.5 * 100
```
So 136.6%.
:::

::: {.panel-tabset}
## 7.18
Is there a significant excess or deficit of total cancer deaths among these workers?

## Solution
We use the poisson test. Let $X$ be the number of cancer deaths. Then we assume $X \sim \mathrm{Pois}(\mu)$. We are testing $H_0: \mu = 20.5$ versus $H_A: \mu > 20.5$. The $p$-value of the exact test is just $P(X \geq 28)$ assuming $X \sim \mathrm{Pois}(20.5)$, since that would be the probability of seeing values as or more extreme than what we saw:
```{r}
#| echo: false
plt_pois(mu = 20.5, lb = 28)
```
```{r}
1 - ppois(q = 27, lambda = 20.5)
```
You can also get this with `poisson.test()`
```{r}
poisson.test(x = 28, alternative = "greater", r = 20.5) |>
  tidy() |>
  select(p.value)
```
The $p$-value is somewhat small. So we perhaps have some weak evidence of larger cancer rates. But the evidence is very weak.
:::

In the same group of workers, 7 deaths due to leukemia are noted. Only 4.5 are expected from statewide rates.

::: {.panel-tabset}
## 7.19
What is the estimated SMR for leukemia?

## Solution
SMR is
$$
\frac{\text{observed}}{\text{expected}}\times 100\% = \frac{7}{4.5}\times 100\%
$$
```{r}
7 / 4.5 * 100
```
So 155.6%.
:::

::: {.panel-tabset}
## 7.20 
Is there a significant excess or deficit of leukemia deaths among these workers?

## Solution
Let $Y$ be the number of leukemia deaths. Then we assume $Y \sim \mathrm{Pois}(\mu)$. We are testing $H_0: \mu = 4.5$ versus $H_A: \mu \neq 4.5$. We do an exact Poisson test here, summing over all probabilities less (or as) likely than the observed value of 7.
```{r}
#| echo: false
plt_pois(mu = 4.5, valmax = 7, rng = c(0, 12)) +
  geom_hline(yintercept = dpois(x = 7, lambda = 4.5), lty = 2)
```
This is done via `poisson.test()`
```{r}
poisson.test(x = 7, r = 4.5) |>
  tidy() |>
  select(p.value)
```
Since the $p$-value is pretty large, we do not have evidence that the leukemia rates differ in this group.

:::{.callout-tip}
## Note
Not for the exam, but for your own edification. The below shows that R's default way is what we discussed:
```{r}
x <- dpois(x = 0:20, lambda = 4.5)
sum(x[x <= x[[8]]])
```
Rosner does 2 times the larger one-tailed $p$-value:
```{r}
2 * (1 - ppois(q = 6, lambda = 4.5))
```
So if you read his solution, the $p$-value will differ.
:::
:::

## Infectious Disease

Suppose the annual incidence of diarrhea (defined as 1+ episodes per year) in a Third-World country is 5% in children under the age of 2.

::: {.panel-tabset}
## 7.21 
If 10 children out of 108 under the age of 2 in a poor rural community in the country have 1+ episodes of diarrhea in a year, then test if this represents a significant departure from the overall rate for the country using the critical-value method.

## Solution
Let $X$ be the number with 1+ episodes in a year. Then $X \sim \mathrm{Binom}(108, p)$. We are told that $\hat{p} = 10 / 108 = 0.09259$. We are testing $H_0: p = 0.05$ versus $H_A: p \neq 0.05$. We can use the normal method since $np(1-p)\geq 5$
```{r}
108 * 0.05 * (1 - 0.05)
```
We calculate the $z$-statistic
$$
z = \frac{\hat{p} - p_0}{\sqrt{p_0(1-p_0)/n}} = \frac{0.09259 - 0.05}{\sqrt{0.05(1-0.05)/108}}
$$
Numerically, the $z$-value is
```{r}
(0.09259 - 0.05) / sqrt(0.05 * (1 - 0.05) / 108)
```
For the critical-value method, we compare this $z$-statistic to $z_{1 - \alpha/2}$. Let $\alpha = 0.05$
```{r}
qnorm(1 - 0.05 / 2)
```
Since $|z| > z_{1-\alpha}$ (2.031 > 1.96), we reject $H_0$ and conclude that diarrhea rate differs from the country as a whole.
:::

::: {.panel-tabset}
## 7.22 
Report a p-value corresponding to your answer to Problem 7.21

## Solution
We could alternatively calculate $P(Z > |z|)$ given $Z \sim N(0,1)$.
```{r}
#| echo: false
plt_t(lb = 2.031, two_sided = TRUE) +
  geom_vline(xintercept = 2.031, lty = 2)
```
Numerically, the $p$-value is
```{r}
2 * pnorm(q = -2.031)
```
Since this is less than 0.05, we again reject at significance level 0.05 and conclude that the diarrhea rate differs than the country as a whole.

You can also use `prop.test()` to do this
```{r}
prop.test(x = 10, n = 108, p = 0.05) |>
  tidy() |>
  select(p.value)
```
Recall that `prop.test()` uses a slightly different method, which is why the $p$-value differs. If you were a 0.05 absolutist, you would be very worried now. But we "flexitarians" are more chill. There is practically no difference between 0.04 and 0.07. Both are pretty weak evidence.
:::

## Ophthalmology

Suppose the distribution of systolic blood pressure in the general population is normal with a mean of 130 mm Hg and a standard deviation of 20 mm Hg. In a special subgroup of 85 people with glaucoma, we find that the mean systolic blood pressure is 135 mm Hg with a standard deviation of 22 mm Hg.

::: {.panel-tabset}
## 7.23 
Assuming that the standard deviation of the glaucoma patients is the same as that of the general population, test for an association between glaucoma and high blood pressure.

## Solution
Let $X_i$ be the systolic blood pressure for glaucoma patient $i$. Then we assume $X \sim N(\mu, \sigma^2)$. We want to test $H_0:\mu = 130$ versus $H_A: \mu \neq 130$. We are told that $\bar{x} = 135$ and $n = 85$. We are further told to assume that we know $\sigma = 20$. We calculate a $z$-statistic
$$
z = \frac{\bar{x} - \mu_0}{\sigma/\sqrt{n}} = \frac{135 - 130}{20 / \sqrt{85}}
$$
Numerically
```{r}
(135 - 130) / (20 / sqrt(85))
```
We compare this to a $N(0,1)$ distribution
```{r}
#| echo: false
plt_t(lb = 2.305, two_sided = TRUE) +
  geom_vline(xintercept = 2.305, lty = 2)
```
```{r}
2 * pnorm(-2.305)
```
So we have some evidence that the systolic blood pressure differs in the glaucoma group than the general population.
:::

::: {.panel-tabset}
## 7.24 
Answer Problem 7.23 without making the assumption concerning the standard deviation.

## Solution
We are in the same set up as in 7.23, except we are told that $s = 22$ and we cannot assume that we know $\sigma$. We calculate a $t$-statistic
$$
t = \frac{\bar{x} - \mu_0}{s/\sqrt{n}} = \frac{135 - 130}{22 / \sqrt{85}}
$$
Numerically
```{r}
(135 - 130) / (22 / sqrt(85))
```
We compare this to a $t_{84}$ distribution
```{r}
#| echo: false
plt_t(lb = 2.095, two_sided = TRUE, df = 84) +
  geom_vline(xintercept = 2.095, lty = 2)
```
```{r}
2 * pt(q = -2.095, df = 84)
```
Again, we have evidence that the glaucoma group has a different SBP than the population as a whole.
:::

## Occupational Health 2

::: {.panel-tabset}
## 7.25 
Suppose it is known that the average life expectancy of a 50-year-old man in 1945 was 18.5 years. Twenty men aged 50 who have been working for at least 20 years in a potentially hazardous industry were ascertained in 1945. On follow-up in 1985 all the men have died, with an average lifetime of 16.2 years and a standard deviation of 7.3 years since 1945. Assuming that life expectancy of 50-year-old men is approximately normally distributed, test if the underlying life expectancy for workers in this industry is shorter than for comparably aged men in the general population.

## Solution
Let $X_i$ be the lifetime of worker $i$. Then we assume that $X_i \sim N(\mu, \sigma^2)$. We are testing $H_0: \mu = 18.5$ versus $H_A: \mu < 18.5$. We are told that $\bar{x} = 16.2$, $s = 7.3$, and $n = 20$. We calculate a $t$-statistic
$$
t = \frac{\bar{x} - \mu_0}{s/\sqrt{n}} = \frac{16.2 - 18.5}{7.3 / \sqrt{20}}
$$
Numerically
```{r}
(16.2 - 18.5) / (7.3 / sqrt(20))
```
We compare this to a $t_{19}$ distribution
```{r}
#| echo: false
plt_t(ub = -1.409, df = 19) +
  geom_vline(xintercept = -1.409, lty = 2)
```
```{r}
pt(q = -1.409, df = 19)
```
The $p$-value is pretty large, so we have at best weak evidence that the average lifetime is less than the general population.
:::


## Cancer 1

::: {.panel-tabset}
## 7.26
An area of current interest in cancer epidemiology is the possible role of oral contraceptives (OC’s) in the development of breast cancer. Suppose that in a group of 1000 premenopausal women ages 40–49 who are current users of OC’s, 15 subsequently develop breast cancer over the next 5 years. If the expected 5-year incidence rate of breast cancer in this group is 1.2% based on national incidence rates, then test the hypothesis that there is an association between current OC use and the subsequent development of breast cancer.

## Solution
Let $X$ be the number with breast cancer after 5 years. Then $X \sim \mathrm{Binom}(1000, p)$. We want to test $H_0: p = 0.012$ versus $H_A: p \neq 0.012$. We are told that $x = 15$ was observed, so $\hat{p} = 15/1000 = 0.015$. We can use a normal approximation since $np_0(1-p_0) > 5$
```{r}
1000 * 0.012 * (1 - 0.012)
```
We calculate a $z$-statistic
$$
z = \frac{\hat{p} - p_0}{\sqrt{p_0(1-p_0)/n}} = \frac{0.015 - 0.012}{0.012(1-0.012)/1000}
$$
Numerically
```{r}
(0.015 - 0.012) / sqrt(0.012 * (1 - 0.012) / 1000)
```
We compare this to a $N(0,1)$ distribution
```{r}
#| echo: false
plt_t(lb = 0.8713, two_sided = TRUE) +
  geom_vline(xintercept = 0.8713, lty = 2)
```
```{r}
2 * pnorm(-0.8713)
```
Since the $p$-value is so large, we have no evidence that these women have a different breast cancer rate than the general population.

In R, we would use `prop.test()`
```{r}
prop.test(x = 15, n = 1000, p = 0.012) |>
  tidy() |>
  select(p.value)
```
Again, the $p$-value differs because R uses a different (better) method.
:::

## Cancer 2

A group of investigators wishes to explore the relationship between the use of hair dyes and the development of breast cancer in females. A group of 1000 beauticians 30–39 years of age is identified and followed for 5 years. After 5 years, 20 new cases of breast cancer have occurred. Assume that breast-cancer incidence over this time period for an average woman in this age group is 7/1000. We wish to test the hypothesis that using hair dyes increases the risk of breast cancer.

::: {.panel-tabset}
## 7.27
Is a one-sided or two-sided test appropriate here?

## Solution
One-sided. We are interested in one direction (hair dies increase the risk).

Let $X$ be the number of breast cancer cases. Then $X \sim \mathrm{Binom}(1000, p)$. We are testing $H_0: p = 7/1000$ versus $H_A: p > 7/1000$.
:::

::: {.panel-tabset}
## 7.28
Test the hypothesis.

## Solution
Since $np_0(1-p_0) \geq 5$, we can use the normal approximation.
```{r}
1000 * 0.007 * (1 - 0.007)
```
With $\hat{p} = 20 / 1000$, we calculate a $z$-statistic
$$
z = \frac{\hat{p} - p_0}{\sqrt{p_0(1-p_0)/n}} = \frac{0.02 - 0.007}{0.007(1-0.007)/1000}
$$
Numerically
```{r}
(0.02 - 0.007) / sqrt(0.007 * (1 - 0.007) / 1000)
```
We compare this to a $N(0,1)$ distribution
```{r}
#| echo: false
plt_t(lb = 4.931, rng = c(-6, 6)) +
  geom_vline(xintercept = 4.931, lty = 2)
```
```{r}
1 - pnorm(4.931)
```
Since the $p$-value is so small, we have strong evidence that these women have a higher breast cancer rate than the general population.

In R, we would use `prop.test()`
```{r}
prop.test(x = 20, n = 1000, p = 0.007, alternative = "greater") |>
  tidy() |>
  select(p.value)
```
Again, the $p$-value differs because R uses a different (better) method.
:::


## Renal Disease 1

The level of serum creatinine in the blood is considered a good indicator of the presence or absence of kidney disease. Normal people generally have low concentrations of serum creatinine, whereas diseased people have high concentrations. Suppose we want to look at the relation between analgesic abuse and kidney disorder. In particular, suppose we look at 15 people working in a factory who are known to be "analgesic abusers" (i.e., they take more than 10 pills per day) and we measure their creatinine levels. The creatinine levels are
```{r}
cret <- tibble(
  creatine = c(0.9, 1.1, 1.6, 2.0, 0.8,
               0.7, 1.4, 1.2, 1.5, 0.8,
               1.0, 1.1, 1.4, 2.2, 1.4)
)
cret
```

::: {.panel-tabset}
## 7.29 
If we assume that creatinine levels for normal people are normally distributed with mean 1.0, then can we make any comment about the levels for analgesic abusers via some statistical test?

## Solution
Let $X_i$ be the creatine level for analgesic abuser $i$. Then we assume that $X_i \sim N(\mu, \sigma^2)$. We want to test $H_0: \mu = 1$ versus $H_A: \mu \neq 1$. We calculate summary statistics
```{r}
cret |>
  summarize(
    mean = mean(creatine),
    sd = sd(creatine),
    n = n())
```
We calculate a $t$-statistic
$$
t = \frac{\bar{x} - \mu_0}{s/\sqrt{n}} = \frac{1.273 - 1}{0.435 / \sqrt{15}}
$$
Numerically
```{r}
(1.273 - 1) / (0.435 / sqrt(15))
```
We compare this to a $t_{14}$ distribution
```{r}
plt_t(lb = 2.431, two_sided = TRUE, df = 14) +
  geom_vline(xintercept = 2.431, lty = 2)
```
```{r}
2 * pt(-2.431, df = 14)
```
Since the $p$-value is so small, we have evidence that the mean creatine levels of the analgesic abusers is different than normal folks.

The real-way in R uses `t.test()`
```{r}
t.test(creatine ~ 1, data = cret, mu = 1) |>
  tidy() |>
  select(p.value)
```
The difference is caused by rounding error when calculating the $t$-statistic.
```{r}
#| echo: false
#| eval: false
cret |>
  summarize(
    mean = mean(creatine),
    sd = sd(creatine),
    n = n()) |>
  mutate(t = (mean - 1) / (sd / sqrt(n))) ->
  temp
2 * pt(-temp$t, df = 14)
```
:::

## Epidemiology

Height and weight are often used in epidemiological studies as possible predictors of disease outcomes. If the people in the study are assessed in a clinic, then heights and weights are usually measured directly. However, if the people are interviewed at home or by mail, then a person’s self-reported height and weight are often used instead. Suppose we conduct a study on 10 people to test the comparability of these two methods. The data for weight are given in the following data frame.

```{r}
wt <- tibble(
  Subject_number = 1:10,
  Self_reported_weight = c(120, 120, 135, 118, 120, 190, 124, 175, 133, 125),
  Measured_weight = c(125, 118, 139, 120, 125, 198, 128, 176, 131, 125),
  Difference = c(-5, 2, -4, -2, -5, -8, -4, -1, 2, 0)
)
wt
```


::: {.panel-tabset}
## 7.33
Should a one-sided or two-sided test be used here?

## Solution
:::

::: {.panel-tabset}
## 7.34
Which test procedure should be used to test the preceding hypothesis?

## Solution
:::

::: {.panel-tabset}
## 7.35 
Conduct the test in Problem 7.34 using the critical-value method with α = .05.

## Solution
:::

::: {.panel-tabset}
## 7.36
Compute the p-value for the test in Problem 7.34.

## Solution
:::

::: {.panel-tabset}
## 7.37
Is there evidence of digit preference among the self- reported weights? Specifically, compare the observed proportion of self-reported weights whose last digit is 0 or 5 with the expected proportion based on chance and report a p-value.

## Solution
:::

## Hypertension

Several studies have been performed relating urinary potassium excretion to blood-pressure level. These studies have tended to show an inverse relationship between these two variables, with the higher the level of potassium excretion, the lower the BP level. Therefore, a treatment trial is planned to look at the effect of potassium intake in the form of supplement capsules on changes in DBP level. Suppose that in a pilot study, 20 people are given potassium supplements for 1 month. The data are as follows:

```{r}
#| echo: false
#| eval: true
#| message: false

# Create the data frame
data <- data.frame(
  Metric = c("Mean change (1 month-baseline)", "sd change", "n"),
  Value = c(-3.2, 8.5, 20)
)

gt(data)
```

::: {.panel-tabset}
## 7.38
What test should be used to assess if potassium supplements have any effect on DBP level?

## Solution
:::

::: {.panel-tabset}
## 7.39 
Perform the test in Problem 7.38 using a two-sided test and report the p-value.

## Solution
:::

::: {.panel-tabset}
## 7.40
Derive a 95% CI for the true mean change based on the preceding data. What is the relationship of your results here and in Problem 7.39?

## Solution
:::

::: {.panel-tabset}
## 7.41
How many subjects need to be studied to have an 80% chance of detecting a significant treatment effect using a two-sided test with an α level of .05 if the mean and sd of the pilot study are assumed to be the population mean and sd?

## Solution
:::

## Hospital Epidemiology

A study was conducted to identify characteristics that would predict 1-year survival for patients admitted to the medical service at New York Hospital [1]. One factor that was considered was the physician’s estimate of the patients’ severity of illness at the time of admission. Suppose that it is expected, based on previous studies in this hospital, that 2/3 of admitted patients will survive for at least 1 year.

::: {.panel-tabset}
## 7.42 
If 47% of 136 patients deemed severely ill survive at least 1 year, then what test can be used to test the hypothesis that the severity-of-illness rating is predictive of 1-year mortality?

## Solution
:::

::: {.panel-tabset}
## 7.43 
Perform the test in Problem 7.42 using the critical-value method based on an α level pf .05.

## Solution
:::

::: {.panel-tabset}
## 7.44 
Provide a 95% CI for the 1-year survival rate among severely ill patients.

## Solution
:::

Suppose that it is expected, based on previous studies, that among all patients who survive hospitalization, 75% will survive for 1 year. Furthermore, of the 136 severely ill patients, 33 die during hospitalization, and an additional 39 die during the 1st year, but after hospitalization.

::: {.panel-tabset}
## 7.45
Test the hypothesis that severity of illness is predictive of 1-year mortality among patients who are discharged from the hospital. Report a p-value.

## Solution
:::

## Cancer 3

Radiotherapy is a common treatment for breast cancer, with about 25% of women receiving this form of treatment. Assume that the figure 25% is based on a very large sample and is known without error. One hypothesis is that radiotherapy applied to the contralateral breast may be a risk factor for development of breast cancer in the opposite breast 5 or more years after the initial cancer. Suppose that 655 women are identified who developed breast cancer in the opposite breast 5 or more years after the initial cancer [2].

::: {.panel-tabset}
## 7.46
If 206 of the women received radiotherapy after their initial diagnosis, then test the hypothesis that radiotherapy is associated with the development of breast cancer in the opposite breast. Please report a p-value.

## Solution
:::

::: {.panel-tabset}
## 7.47
Provide a 95% CI for the true proportion of women with contralateral breast cancer who received radiotherapy treatment.

## Solution
:::

::: {.panel-tabset}
## 7.48
Suppose that our p-value in Problem 7.46 = .03 (this is not necessarily the actual p-value in Problem 7.46). If we conduct a test using the critical-value method with α = .05, then would we accept or reject $H_0$ and why? (Do not actually conduct the test.)

## Solution
:::

::: {.panel-tabset}
## 7.49
Suppose we wanted a 99% CI instead of a 95% CI in Problem 7.47. Would the length of the 99% CI be narrower, wider, or the same as the CI in Problem 7.47? (Do not actually construct the interval.)

## Solution
:::


## Renal Disease 2

To compare two methods for the determination of uric acid, 23 blood samples from 23 individuals were divided and analyzed both by the colorimetric method and the uricase method. Suppose the sample means of the colorimetric and uricase assessments were 6.26 and 6.20 mg/dL, respectively, the sample standard deviation of the paired difference between repeated assessments is 0.50 mg/dL, and it is reasonable to assume that this paired difference has a normal distribution.

::: {.panel-tabset}
## 7.50 
What is the best estimate for the mean difference between approaches?

## Solution
:::

::: {.panel-tabset}
## 7.51 
Construct a 95% CI for the mean difference.

## Solution
:::

::: {.panel-tabset}
## 7.52 
What can be concluded about the difference between assessments by the two methods?

## Solution
:::


## Cardiovascular Disease

The relationship between serum cholesterol and coronary-heart-disease mortality has been a subject of much debate over the past 30 years. Some data relevant to this question were presented in the Whitehall study. It was shown there that among 3615 men 40–64 years of age who were in the top quintile of serum cholesterol at baseline, 194 died from coronary heart disease over the next 10 years. Suppose the incidence rate of coronary-heart-disease mortality over 10 years among men in this age group in Great Britain is 4%.

::: {.panel-tabset}
## 7.53
What test can be performed to compare the incidence rate of coronary-heart-disease mortality among men in the top cholesterol quintile vs general-population incidence rates?

## Solution
:::

::: {.panel-tabset}
## 7.54
Implement the test in Problem 7.53 and report a p-value.

## Solution
:::

::: {.panel-tabset}
## 7.55 
Construct a 95% CI for the true incidence rate in the group of 40–64-year-old men in the top quintile of serum cholesterol.

## Solution
:::

## Cancer 4

It is well established that exposure to ionizing radiation at or after puberty increases a woman’s risk of breast cancer. However, it is uncertain whether such exposure early in life is also carcinogenic. A study was performed in a cohort of 1201 women who received x-ray treatment in infancy for an enlarged thymus gland and were followed prospectively for 36 years. It was found that 22 breast cancers occurred over a 36-year period among the 1201 women.

::: {.panel-tabset}
## 7.56
If the expected incidence rate of breast cancer in this age group over this time period is 1 event per 200 women based on New York State cancer-incidence rates, then test the hypothesis that irradiation has an effect on breast-cancer incidence. Please report a p-value and construct a 95% CI for the incidence rate among the exposed group.

## Solution
:::

It was also found that 6 breast cancers occurred over a 36-year period among a subgroup of 138 women who were exposed to a high radioactive dose (0.50–1.99 gray, where a gray is a unit of radiation absorbed by the breast tissue).

::: {.panel-tabset}
## 7.57 
Test the hypothesis that the high-risk subgroup is at excess risk for breast cancer, and report a p-value.

## Solution
:::
