---
title: "Chapter 12: ANOVA"
author: "David Gerard"
date: today
bibliography: "../bib.bib"
---

```{r}
#| message: false
#| echo: false
library(tidyverse)
library(broom)
library(patchwork)
knitr::opts_chunk$set(echo = TRUE, 
            fig.width = 4, 
            fig.height = 3, 
            fig.align = "center")
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

# Learning Objectives

- ANOVA
- Chapter 12 of Rosner

# Interpreting Models

- ANOVA is just multiple linear regression with categorical explanatory variable(s).

- ANCOVA is just multiple linear regression with a categorical explanatory variable of interest and a few other explanatory variables, at least one of which is continuous.

- There are really only two differences:
  1. The notation.
  2. More emphasis on reporting sums of squares.
  
# Interpreting One-way ANOVA Models

- The Swiss Analgesic Study aimed to evaluate the impact of phenacetin-containing analgesics on kidney function and health.

It involved 624 women from Basel, Switzerland, who had high phenacetin intake and 626 women with low or no phenacetin intake (control group). The study used urine N-acetyl-P-aminophenyl (NAPAP) levels to measure recent phenacetin use, dividing the study group into high-NAPAP and low-NAPAP subgroups. Both subgroups had higher NAPAP levels than the control group. The women were examined in 1967-1968 and again in 1969, 1970, 1971, 1972, 1975, and 1978, with kidney function assessed through various laboratory tests, including serum-creatinine levels. You can read about the data [here](https://dcgerard.github.io/stat_320/data.html#swiss), and load it in via:
```{r}
#| message: false
swiss <- read_csv("https://dcgerard.github.io/stat_320/data/swiss.csv")
```


# Sums of Squares for One-way ANOVA Models

# Interpreting Two-way ANOVA Models

# Sums of Squares for Two-way ANOVA models

# Interpreting ANCOVA Models

# Sums of Squares for ANCOVA Models

# References
