---
title: "Power Calculations"
author: "David Gerard"
format: html
date: today
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
            fig.width = 4, 
            fig.height = 3, 
            fig.align = "center")
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

-   For $t$-methods, you use `power.t.test()` to calculate do power and smaple size calculations. It takes as input four of the following:
    - `n`: The sample size $n$
    - `delta`: The effect size (difference in means) $|\mu_1 - \mu_0|$
    - `sd`: The standard deviation of the data $\sigma$
    - `sig.level`: The signficicance level $\alpha$
    - `power`: The power $1 - \beta$.
    
- You must put values for exactly four of the above. The fifth should be `NULL` and the function will return the fifth value.
    
-   Other inputs are for the type of test:
    - `type`: Use `"one.sample"` for one-sample $t$-tests and `"two.sample"` for two-sample $t$-tests.
    - `alternative`: Either `"two.sided"` or `"one.sided"`. 
    
-   Suppose we plan on running a study with 100 participants of low socioeconomic status (SES). The mean birthweight in america is 120 oz. A pilot study suggested that the average birthweight of low SES babies is 115 oz with a standard deviation of 24 oz. What is the power of a test with a significance level of 0.05?

    ```{r}
    power.t.test(
      n = 100, 
      delta = 5, 
      sd = 24, 
      sig.level = 0.05, 
      type = "one.sample", 
      alternative = "one.sided")
    ```

-   Let's compare that power to the $z$-test calculation from Rosner

    ```{r}
    pnorm(qnorm(0.05) + sqrt(100) * 5 / 24)
    ```

- It's a little different because Rosner uses $z$-methods instead of $t$-methods for power and sample size calculations. But it's not too different to be practically important, especially since power and sample size calculations are mostly just ~~wild~~ educated guesses.


::: {.panel-tabset}
## Exercise
What sample size would be needed for a power of at least 0.8?

## Solution
```{r}
power.t.test(
  power = 0.8, 
  delta = 5, 
  sd = 24, 
  sig.level = 0.05, 
  type = "one.sample", 
  alternative = "one.sided")
```

We need a sample size of at least 144.
:::

::: {.panel-tabset}  
## Exercise
A new drug is proposed to prevent glaucoma among people with high intraocular pressure (IOP). A pilot study is conducted with 10 individuals. After 1 month of using the drug, their IOP decreases by 5 mm HG with a standard deviation of 10 mm HG. What is the sample size needed to achieve 90% power for a *two-sided* test with significance level of 0.05.

## Solution
```{r}
power.t.test(
  delta = 5, 
  sd = 10,
  sig.level = 0.05,
  power = 0.9,
  type = "one.sample",
  alternative = "two.sided")
```

We need a smaple size of at least 44.
:::
