---
title: "Summary Statistics in R"
author: "David Gerard"
date: "`r Sys.Date()`"
---

- We'll use the lead data as an example. Read about it [here](../data.qmd#lead).

  ```{r}
  #| message: false
  library(tidyverse)
  lead <- read_csv("https://dcgerard.github.io/stat_320/data/lead.csv")
  ```

- You calculate the summary statistics (mean/median/quantiles/variance/standard deviation) all within a `summarize()` call.

  ```{r}
  summarize(
    lead, 
    Mean = mean(iqf, na.rm = TRUE), 
    Min = min(iqf, na.rm = TRUE),
    Q25 = quantile(iqf, probs = 0.25, na.rm = TRUE),
    Med = median(iqf, na.rm = TRUE), 
    Q75 = quantile(iqf, probs = 0.75, na.rm = TRUE),
    Max = max(iqf, na.rm = TRUE),
    Var = var(iqf, na.rm = TRUE),
    SD = sd(iqf, na.rm = TRUE)
  )
  ```

- The values on the left of `=` are the names of the summaries and are up to you.

- The values on the right of `=` are the function calls for the summaries.
  - `mean()`: the arithmetic mean.
  - `min()`: the minimum. Same as `quantile(x, probs = 0)`
  - `quantile()`: the quantiles. You specify which quantile with the `probs` argument.
  - `median()`: the median. Same as `quantile(x, probs = 0.5)`
  - `max()`: the maximum. Same as `quantile(x, probs = 1)`
  - `var()`: the sample variance.
  - `sd()`: the sample standard deviation.

- I have the `na.rm = TRUE` argument because there are some children who did not have a `iqf` score. These are "missing" and encoded with `NA`. If you do not provide that argument, R doesn't know what those values are and so returns an `NA` or errors.
  ```{r}
  summarize(
    lead, 
    Mean = mean(iqf), 
    Min = min(iqf),
    # Q25 = quantile(iqf, probs = 0.25), # errors
    Med = median(iqf), 
    # Q75 = quantile(iqf, probs = 0.75), # errors
    Max = max(iqf),
    Var = var(iqf),
    SD = sd(iqf)
  )
  ```


- You can also apply these functions on vectors that you extract from the data frame.

  ```{r}
  var(lead$iqf, na.rm = TRUE)
  ```
 
- Let's demonstrate some properties. Variance is invariant to shift
  ```{r}
  var(lead$iqf + 10000, na.rm = TRUE)
  ```
  but scales with the square of the multiplicative factor
  ```{r}
  var(10 * lead$iqf, na.rm = TRUE)
  10^2 * var(lead$iqf, na.rm = TRUE)
  ```

- The standard deviation scales with the multiplicative factor because it is the square root of the variance.
  ```{r}
  sd(10 * lead$iqf, na.rm = TRUE)
  10 * sd(lead$iqf, na.rm = TRUE)
  ```

- The mean and quantiles shift and scale with the additive and multiplicative factors.
  ```{r}
  mean(lead$iqf * 10 + 20, na.rm = TRUE)
  10 * mean(lead$iqf, na.rm = TRUE) + 20
  
  quantile(lead$iqf * 10 + 20, probs = c(0.25, 0.5, 0.75), na.rm = TRUE)
  10 * quantile(lead$iqf, probs = c(0.25, 0.5, 0.75), na.rm = TRUE) + 20
  ```

- **Exercise**: Calculate the mean and median of the [birthweight data](../data.qmd#birthweight). What is the more appropriate measure of center?

  ```{r}
  #| echo: false
  #| eval: false
  birthweight <- read_csv("https://dcgerard.github.io/stat_320/data/birthweight.csv")
  ggplot(birthweight, aes(x = weight)) +
    geom_histogram()
  ## looks skewed left, so medium?
  summarize(
    birthweight,
    Mean = mean(weight),
    Med = median(weight)
  )
  ## But is also bimodal, so maybe need multiple measures of center (for premies and normals).
  ```

- You can calculate grouped summaries (a summary for each group) by grouping the data first.
  ```{r}
  lead |>
    group_by(Group) |>
    summarize(
      Mean = mean(iqf, na.rm = TRUE),
      SD = sd(iqf, na.rm = TRUE)
    )
  ```

- Group summaries are where the power of descriptive statistics really comes into play. Here, we see that the exposed group has a lower IQ on average than the control group. Whether this is real signal will have to be answered via a formal hypothesis test. But the descriptive statistics gives us some initial information on the data.

- **Exercise**: What about different lead groups? Calculate descriptive statistics for the different lead groups.

  ```{r}
  #| echo: false
  #| eval: false
  lead |>
    group_by(lead_grp) |>
    summarize(
      Mean = mean(iqf, na.rm = TRUE),
      SD = sd(iqf, na.rm = TRUE)
    )
  ```

 
