---
title: "Chapter 2 Notes: Descriptive Statistics"
author: "David Gerard"
format: html
date: today
---

## Measures of Location

- Observe $X_1, X_2, \dots, X_n$
  - Sample of numeric values
  - Subscript indexes the units

- Example: $X_i =$ Birthweight for baby $i$

- Measure of location = center of a sample (statistic) or a population (parameter)

- Arithmetic **Mean**
  $$
  \bar{X} = \frac{1}{n} \sum_{i=1}^n X_i = \frac{1}{n} (X_1 + X_2 + \dots + X_n)
  $$

- Example: $X_1 = 2$, $X_2 = 5$, $X_3 = -4$
    $$
    \sum_{i=1}^3 X_i = X_1 + X_2 + X_3 = 2 + 5 + (-4)
    $$
    
    $$
    \sum_{i=1}^2 X_i = X_1 + X_2 = 2 + 5
    $$
    
    $$
    \sum_{i=2}^2 X_i = X_2 = 5
    $$
    
    $$
    \bar{X} = \frac{1}{3} \sum_{i=1}^3 X_i = \frac{1}{3} (2 + 5 - 4) = \frac{3}{3} = 1
    $$

- $\bar{X}$ is sensitive to extreme observations.

- Example with extreme value:

    $$
    X_4 = 3997
    $$
    
    $$
    \frac{1}{4} \sum_{i=1}^4 X_i = \frac{1}{4} (2 + 5 - 4 + 3997) = \frac{4000}{4} = 1000
    $$

- **Median**
  - For $n$ odd $\Rightarrow \frac{n+1}{2}$ th largest observation.
  - For $n$ even $\Rightarrow$ Average of $\frac{n}{2}$ th and $\frac{n}{2} + 1$ th largest observations.

- Example: $X_1 = 2$, $X_2 = 5$, $X_3 = -4 \Rightarrow -4, 2, 5$

    $$
    \text{Median}(X) = 2
    $$

- If $X_4 = 3997$

    $$
    \text{Median}(X) = \frac{2 + 5}{2} = 3.5
    $$

- If distribution is symmetric, $\text{median}(X) \approx \bar{X}$.

- Mean follows the skew of distribution (dashed is median, dotted is mean):

```{r}
#| echo: false
#| eval: true
#| message: false
library(ggplot2)

# Right skew plot
ggplot(data.frame(x = c(-2, 6)), aes(x = x)) +
  stat_function(fun = dlnorm, args = list(meanlog = 0.5, sdlog = 0.8), geom = "area", fill = "skyblue", alpha = 0.5) +
  geom_vline(xintercept = exp(0.5), linetype = "dashed") +
  geom_vline(xintercept = exp(0.5 + 0.8^2/2), linetype = "dotted") +
  labs(title = "Right Skew", x = "Years of oral conception use", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_blank())

# Left skew plot
ggplot(data.frame(x = c(-2, 6)), aes(x = x)) +
  stat_function(fun = dlnorm, args = list(meanlog = -0.5, sdlog = 0.8), geom = "area", fill = "pink", alpha = 0.5) +
  geom_vline(xintercept = exp(-0.5), linetype = "dashed") +
  geom_vline(xintercept = exp(-0.5 + 0.8^2/2), linetype = "dotted") +
  labs(title = "Left Skew", x = "Relative Humidity in Summer (Ohio)", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  scale_x_reverse()
```

- Use mean if total is important.

- Use median if lots of skew.

- A **mode** is a frequently occurring value.

- Types of Modalities:
  - **Unimodal**: Example - male height
  - **Bimodal**: Example - male and female height
  - **Trimodal** etc...

  ```{r}
  #| echo: false
  #| eval: true
  #| message: false
  library(ggplot2)
  
  # Unimodal plot
  ggplot(data.frame(x = c(-3, 3)), aes(x = x)) +
    stat_function(fun = dnorm, args = list(mean = 0, sd = 0.5), geom = "area", fill = "lightblue") +
    labs(title = "Unimodal Distribution", x = "Male Height", y = "Count") +
    theme_minimal()
  
  # Bimodal plot
  ggplot(data.frame(x = c(-3, 3)), aes(x = x)) +
    stat_function(fun = dnorm, args = list(mean = -1, sd = 0.5), geom = "area", fill = "lightgreen") +
    stat_function(fun = dnorm, args = list(mean = 1, sd = 0.5), geom = "area", fill = "lightgreen") +
    labs(title = "Bimodal Distribution", x = "Male and Female Height", y = "Count") +
    theme_minimal()
  
  # Trimodal plot
  ggplot(data.frame(x = c(-4, 4)), aes(x = x)) +
    stat_function(fun = dnorm, args = list(mean = -2, sd = 0.5), geom = "area", fill = "pink") +
    stat_function(fun = dnorm, args = list(mean = 0, sd = 0.5), geom = "area", fill = "pink") +
    stat_function(fun = dnorm, args = list(mean = 2, sd = 0.5), geom = "area", fill = "pink") +
    labs(title = "Trimodal Distribution", x = "", y = "Count") +
    theme_minimal()
  ```

- The mode is typically not used as a real measure of center but rather as a way to describe distribution.

### Properties of $\bar{X}$

- Suppose you have a frequency table.

- The intervals between menstrual periods (days):

    | Value | Freq |
    |-------|------|
    | 24    | 5    |
    | 25    | 10   |
    | 26    | 28   |
    | 27    | 64   |
    | 28    | 185  |

- Mean of $X$: 

    $$
    n = 5 + 10 + 28 + 64 + 185 = 292
    $$

    $$
    \bar{X} = \frac{1}{n} \sum_{i=1}^n X_i = \frac{1}{292} (5 \cdot 24 + 10 \cdot 25 + 28 \cdot 26 + 64 \cdot 27 + 185 \cdot 28) = 27.42
    $$

-  Median of $X$
    
    $$
    \text{Median}(X) = \frac{146^{\text{th}} \text{ and } 147^{\text{th}} \text{ values}}{2}
    $$
    
    $$
    = \frac{28 + 28}{2} = 28
    $$

- Let $y_i = x_i + C$, then $\bar{y} = \bar{X} + C$

- Proof:

    $$
    \bar{y} = \frac{1}{n} \sum_{i=1}^n (x_i + C) = \frac{1}{n} \sum_{i=1}^n x_i + \frac{1}{n} \sum_{i=1}^n C = \bar{X} + C
    $$

- Example: Let $y_i$ = deviation from 28 days cycle.

    $$
    y_i = x_i - 28
    $$
    
    $$
    \bar{y} = 27.42 - 28 = -0.58
    $$
    
    
- Also true for Median: 
    $$
    \text{Median}(y) = \text{Median}(x) + C
    $$

- Let $y_i = C x_i$, then $\bar{y} = C \bar{X}$

- Proof:

    $$
    \bar{y} = \frac{1}{n} \sum_{i=1}^n C x_i = C \cdot \frac{1}{n} \sum_{i=1}^n x_i = C \bar{X}
    $$

- Example: Change units from days to weeks.
    
    $$
    y_i = \frac{1}{7} x_i
    $$
    
    $$
    \bar{y} = \frac{1}{7} \cdot 27.42 \approx 3.92
    $$

- If $y_i = a x_i + C$, then $\bar{y} = a \bar{X} + C$

**Exercise**: What is the mean menstrual cycle deviation from 4 weeks?

```{block}
#| eval: false
#| echo: false
$$
3.92 - 4 = -0.08
$$
```


### Measures of Spread

- **Spread**: How far apart numbers are.
- **Range**: $\text{Max} - \text{Min}$ (sensitive to extreme values).
- **Inter-quartile Range (IQR)**: $75^{\text{th}}$ percentile - $25^{\text{th}}$ percentile.

- $p^{\text{th}}$ percentile = value $V_p$ such that $p\%$ of points are at or below $V_p$.

- Example: Median = $50^{\text{th}}$ percentile.

- **Quantile**: A unit of proportions instead of percent.
  - $0.75$ quantile = $75^{\text{th}}$ percentile.

- Example: $X_1 = 2$, $X_2 = 5$, $X_3 = -4$
    - $\frac{1}{3}$ quantile = $-4$
    - $\frac{2}{3}$ quantile = $2$
    - $1$ quantile = $5$

- What about the $40^{\text{th}}$ percentile?
    - Use some average of $-4$ and $2$, but definition varies.

- **Variance**: Average of squared deviations.

    $$
    s^2 = \frac{1}{n-1} \sum_{i=1}^n (X_i - \bar{X})^2
    $$

- $n-1$ because we lose some information by estimating $\bar{X}$.

- **Standard Deviation**: Square root of variance.

    $$
    s = \sqrt{\frac{1}{n-1} \sum_{i=1}^n (X_i - \bar{X})^2}
    $$

- Puts this measure of center on the same scale as units (e.g., $oz$ instead of $oz^2$).

- Let $y_i = c_1 x_i + c_2$, then $s^2(y) = c_1^2 s^2(x)$.

- Only scaling affects variance and standard deviation.

    $$
    s(y) = c_1 s(x)
    $$

- Why?

  ```{r}
  #| echo: false
  #| message: false
  #| warning: false
  library(ggplot2)
  
  set.seed(1)
  x <- rnorm(100)
  ggplot(data.frame(x = x), aes(x = x)) +
    geom_histogram(binwidth = 0.5, fill = "lightblue", alpha = 0.7) +
    labs(title = "Original Distribution", x = "X", y = "Count") +
    theme_minimal() +
    xlim(-5, 5)
  
  ggplot(data.frame(x = x + 2), aes(x = x)) +
    geom_histogram(binwidth = 0.5, fill = "lightgreen", alpha = 0.7) +
    labs(title = "Shifted Distribution (X + C2)", x = "X + C2", y = "Count") +
    theme_minimal() +
    xlim(-5, 5)
  
  ggplot(data.frame(x = x * 2), aes(x = x)) +
    geom_histogram(binwidth = 0.5, fill = "lightcoral", alpha = 0.7) +
    labs(title = "Scaled Distribution (C1 * X + C2)", x = "C1 * X + C2", y = "Count") +
    theme_minimal() +
    xlim(-5, 5)
  ```

**Exercise**: What is $s^2(y)$ when $y_i = c_1 x_i$?

```{block}
#| eval: false
#| echo: false
$$
s^2(y) = c_1^2 s^2(x)
$$
```

[R Notebook on Mean / Median / SD / Variance](./02_descriptive.qmd)

