---
title: "Discrete Probability Distributions"
author: "David Gerard"
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
            fig.width = 4, 
            fig.height = 3, 
            fig.align = "center")
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

```{r}
#| message: false
library(tidyverse)
```


# Provided Distribution

If given a probability mass function, can create a data frame of it
```{r}
pmf <- tibble(r = 0:4,
       pr = c(0.008, 0.076, 0.265, 0.411, 0.240)
)
```

We can verify that the PMF sums to 1
```{r}
sum(pmf$pr)
```

We can calculate the mean and variance like so
```{r}
mu <- sum(pmf$r * pmf$pr)
mu
sigma2 <- sum((pmf$r - mu)^2 * pmf$pr)
sigma2
```

You can get the CDF via `cumsum()`
```{r}
pmf |>
  mutate(cdf = cumsum(pr))
```


We can plot it
```{r}
ggplot(pmf, aes(x = r, y = pr)) +
  geom_col(fill = "black")
```


# Binomial Distribution

- The PMF is `dbinom()`. 

  About 60% of all white blood cells are neutrophils. If we observe 10 white blood cells, the probability of seeing 4 neutrophils is
  
  $$
  \binom{10}{4} 0.6^4 0.4^6
  $$
  
  ```{r}
  dbinom(x = 4, size = 10, prob = 0.6)
  ```
  
  ```{r}
  #| echo: false
  n <- 10
  tibble(r = 0:n,
         prob = stats::dbinom(x = 0:n, size = n, prob = 0.6)) |>
    mutate(is_four = r == 4) |>
    ggplot(aes(x = r, y = prob, fill = is_four)) +
    geom_col() +
    scale_fill_manual(values = palette.colors(n = 2, palette = "Okabe-Ito")) +
    scale_x_continuous(breaks = 0:n) +
    guides(fill = "none")
  ```

- The CDF is `pbinom()`:

  $$
  Pr(X \leq x) = \sum_{r=0}^x\binom{10}{r} 0.6^r 0.4^{n-r}
  $$
  ```{r}
  pbinom(q = 4, size = 10, prob = 0.6)
  ```
  
  ```{r}
  #| echo: false
  n <- 10
  tibble(r = 0:n,
         prob = stats::dbinom(x = 0:n, size = n, prob = 0.6)) |>
    mutate(is_four = r <= 4) |>
    ggplot(aes(x = r, y = prob, fill = is_four)) +
    geom_col() +
    scale_fill_manual(values = palette.colors(n = 2, palette = "Okabe-Ito")) +
    scale_x_continuous(breaks = 0:n) +
    guides(fill = "none")
  ```

- The quantile function is `qbinom()`.

  $$
  f(p) = \min(x) \text{ such that } p \leq Pr(X \leq x)
  $$
  
  E.g., the quantile function applied at 0.55 is 6
  ```{r}
  qbinom(p = 0.55, size = 10, prob = 0.6)
  ```
  because the CDF at 6 is above 0.55 and the CDF at 5 is below 0.55.
  ```{r}
  pbinom(q = 5, size = 10, prob = 0.6)
  pbinom(q = 6, size = 10, prob = 0.6)
  ```

- You generate random samples from the binomial distribution with `rbinom()`

  ```{r}
  #| echo: false
  set.seed(3)
  ```


  ```{r}
  x <- rbinom(n = 100, size = 10, prob = 0.6)
  ```
  
  ```{r}
  #| echo: false
  tibble(x = x) |>
    ggplot(aes(x = x)) +
    geom_bar(fill = "black") +
    scale_x_continuous(breaks = 0:n) 
  ```

  ```{r}
  x <- rbinom(n = 10000, size = 10, prob = 0.6)
  ```
  
  ```{r}
  #| echo: false
  tibble(x = x) |>
    ggplot(aes(x = x)) +
    geom_bar(fill = "black") +
    scale_x_continuous(breaks = 0:n) 
  ```

- **Exercise**: The underlying incidence rate of chronic bronchitis in the first year of life is 0.05. What is the probability of obtaining at least 75 cases of chronic bronchitis in the first year of life among 1500 families?

```{r}
#| eval: false
#| echo: false
1 - pbinom(q = 74, size = 1500, prob = 0.05)
pbinom(q = 74, size = 1500, prob = 0.05, lower.tail = FALSE)
```

- **Exercise**: Suppose a group of 100 women ages 60--64 received a new flu vaccine in 2004, and 5 of them died within the next year. Is this event unusual? According to life tables, the probability of death for this age group in the next year is 0.009. Calculate the probability that five or more such women would die under normal circumstances if the flu vaccine had no effect.
  - What is the expected number of such women (out of 100) that we would expect to die in th next year?

```{r}
#| echo: false
#| eval: false
1 - stats::pbinom(q = 4, size = 100, prob = 0.009)
## yes very unusual.
```
  
```{block}
#| eval: false
#| echo: false
100 * 0.009 = 0.9. So about 1 woman out of 100.
```
  

# Poisson Distribution

- The PMF is `dpois()`. 

  Number of deaths from typhoid-fever is over a 1-year period approximately Poisson with rate $\lambda = 4.6$. The probability of exactly 3 deaths is
  
  $$
  e^{-4.6}\frac{4.6^3}{3!}
  $$

  ```{r}
  dpois(x = 3, lambda = 4.6)
  ```
  
  ```{r}
  #| echo: false
  n <- 15
  tibble(k = 0:n,
         prob = stats::dpois(x = 0:n, lambda = 4.6)) |>
    mutate(is_three = k == 3) |>
    ggplot(aes(x = k, y = prob, fill = is_three)) +
    geom_col() +
    scale_fill_manual(values = palette.colors(n = 2, palette = "Okabe-Ito")) +
    scale_x_continuous(breaks = 0:n) +
    guides(fill = "none")
  ```

- The CDF is `ppois()`:

  $$
  Pr(X \leq x) = \sum_{k=0}^{x}e^{-4.6}\frac{4.6^k}{k!}
  $$
  ```{r}
  ppois(q = 3, lambda = 4.6)
  ```

  ```{r}
  #| echo: false
  n <- 15
  tibble(k = 0:n,
         prob = stats::dpois(x = 0:n, lambda = 4.6)) |>
    mutate(is_three = k <= 3) |>
    ggplot(aes(x = k, y = prob, fill = is_three)) +
    geom_col() +
    scale_fill_manual(values = palette.colors(n = 2, palette = "Okabe-Ito")) +
    scale_x_continuous(breaks = 0:n) +
    guides(fill = "none")
  ```

- The quantile function is `qpois()`.

  $$
  f(p) = \min(x) \text{ such that } p \leq Pr(X \leq x)
  $$
  
  E.g., the quantile function applied at 0.55 is 5
  ```{r}
  qpois(p = 0.55, lambda = 4.6)
  ```
  because the CDF at 5 is above 0.55 and the CDF at 4 is below 0.55.
  ```{r}
  ppois(q = 4, lambda = 4.6)
  ppois(q = 5, lambda = 4.6)
  ```

- You generate random samples from the poisson distribution with `rpois()`

  ```{r}
  #| echo: false
  set.seed(3)
  ```


  ```{r}
  x <- rpois(n = 100, lambda = 4.6)
  ```
  
  ```{r}
  #| echo: false
  tibble(x = x) |>
    ggplot(aes(x = x)) +
    geom_bar(fill = "black") +
    scale_x_continuous(breaks = 0:n) 
  ```

  ```{r}
  x <- rpois(n = 10000, lambda = 4.6)
  ```
  
  ```{r}
  #| echo: false
  tibble(x = x) |>
    ggplot(aes(x = x)) +
    geom_bar(fill = "black") +
    scale_x_continuous(breaks = 0:n) 
  ```
  
# Poisson Approximation to Binomial

- For $n$ large, $p$ small, and $np$ intermediate, we have that if $X \sim Binom(n, p)$ then we also have approximately that $X \sim Pois(np)$.

- Rule of thumb: $n \geq 100$ and $p \leq 0.01$

- Example:

  ```{r}
  #| eval: false
  n <- 100
  p <- 0.01
  tibble(
    Binom = dbinom(x = 0:5, size = n, prob = p),
    Pois = dpois(x = 0:5, lambda = n * p)
  )
  ```

  ```{r}
  #| eval: true
  #| echo: false
  n <- 100
  p <- 0.01
  tibble(
    Binom = dbinom(x = 0:5, size = n, prob = p),
    Pois = dpois(x = 0:5, lambda = n * p)
  ) |>
    gt::gt() |>
    gt::fmt_number()
  ```
  
- You don't use this anymore to actually calculate binomial probabilities, since computers do that efficiently without resorting to an approximation.

- This is mostly useful in cases to justify using the Poisson.

- E.g., we see monthly number of cases of Guillain-Barré syndrome in Finland
  - April 1984:  3 
  - May 1984:  7  
  - June 1984:  0  
  - July 1984:  3  
  - August 1984:  4  
  - September 1984:  4  
  - October 1984:  2
  
  The distribution of the number of cases during a month is likely well approximated by a binomial, with $n$ equaling the population of Finland. But we don't know $n$, so we can use a Poisson distribution to model these counts.

