---
title: "Chapter 5 Notes: Continuous Distributions"
author: "David Gerard"
format: html
date: today
---

- A continuous random variable "takes on decimal values."
- For such random variables, the probability at any specific value is $0$.

- **Example:** 
  - $P(\text{a man is exactly 6'2.35792471613"}) \approx 0$?
  - $P(\text{a man is exactly 6'2"}) \approx 0$
  - Men are a little above or a little below.

- However, we know some regions are more likely than others.
  - $P(5' \leq X \leq 7') > P(0 \leq X \leq 1')$

- We describe this intuition with a **PDF** (Probability Density Function).
- A Probability Density Function of a random variable $X$ is a function, $f$, where:
  - $P(a \leq X \leq b) = \text{area below curve between } a \text{ and } b$

- The **CDF** (Cumulative Distribution Function) is again the $F(x) = P(X \leq x)$.

- **Example:** Let $X = \text{Serum triglyceride level}$

  ```{r}
  #| echo: false
  library(ggplot2)
  df <- data.frame(x = seq(0, 200, length.out = 200))
  df$pdf <- dchisq(x = df$x/8, df = 8)
  df2 <- data.frame(x = seq(50, 100, length.out = 200))
  df2$pdf <- dchisq(x = df2$x/8, df = 8)
  ggplot() +
    geom_line(data = df, mapping = aes(x = x, y = pdf)) +
    theme_minimal() +
    xlab("Serum Triglycerides (mg/dL)") +
    ylab("Density") +
    geom_area(data = df2, mapping = aes(x = x, y = pdf), fill = "blue", alpha = 0.5) +
    ggtitle("Pr(50 ≤ X ≤ 100)")
  ```

  ```{r}
  #| echo: false
  library(ggplot2)
  df <- data.frame(x = seq(0, 200, length.out = 200))
  df$pdf <- dchisq(x = df$x/8, df = 8)
  df2 <- data.frame(x = seq(0, 100, length.out = 200))
  df2$pdf <- dchisq(x = df2$x/8, df = 8)
  ggplot() +
    geom_line(data = df, mapping = aes(x = x, y = pdf)) +
    theme_minimal() +
    xlab("Serum Triglycerides (mg/dL)") +
    ylab("Density") +
    geom_area(data = df2, mapping = aes(x = x, y = pdf), fill = "blue", alpha = 0.5) +
    ggtitle("Pr(X ≤ 100)")
  ```

- **Expected value**, $\mu$, the average $X$ over many trials:
  - $\mu = \int_{-\infty}^{\infty} x f(x) \, dx \quad \text{where } f(x) = \text{density}$

- **Variance**: Average squared distance
  - $\sigma^2 = E\left((X - \mu)^2\right) = E(X^2) - \mu^2$
  - $\sigma^2 = \int_{-\infty}^{\infty} (x - \mu)^2 f(x) \, dx$

- Most common continuous distribution: Normal distribution
  - $f(x) = \frac{1}{\sqrt{2 \pi \sigma^2}} \exp\left(-\frac{1}{2} \frac{(x - \mu)^2}{\sigma^2}\right) \quad \text{if } X \sim N(\mu, \sigma^2)$

  - Function of $\sigma^2$ (variance) at $\mu$ (mean)
  
  - Also, if $X \sim N(\mu, \sigma^2)$, then $E(X) = \mu$, $Var(X) = \sigma^2$

- **Normal distribution curve**:
  - A typical bell-shaped curve centered at $\mu$

  ```{r}
  #| echo: false
  #| warning: false
  x_values <- seq(-4, 4, by = 0.01)
  y_values <- dnorm(x_values)
  data <- data.frame(
    x = x_values,
    y = y_values
  )
  ggplot(data, aes(x = x, y = y)) +
    geom_line(size = 1) +
    geom_area(aes(x = ifelse(x >= -3 & x <= 3, x, NA)), fill = "lightblue", alpha = 0.5) +
    geom_vline(xintercept = c(-3, -2, -1, 0, 1, 2, 3), linetype = "dashed") +
    labs(
      x = "x",
      y = "Density",
      title = "Normal Distribution with Mean and Standard Deviations"
    ) +
    scale_x_continuous(breaks = c(-3, -2, -1, 0, 1, 2, 3), labels = c(expression(mu - 3*sigma), expression(mu - 2*sigma), expression(mu - sigma), expression(mu), expression(mu + sigma), expression(mu + 2*sigma), expression(mu + 3*sigma))) +
    theme_minimal()
  ```


- The **standard normal distribution** is $N(0, 1)$.

- **Properties**:
  1. **68-95-99.7 rule**:
     - 68% of area within $\pm 1\sigma$
     - 95% of area within $\pm 2\sigma$
     - 99.7% of area within $\pm 3\sigma$
  2. Symmetric: $f(\mu - x) = f(\mu + x)$
  3. Mean = median = $\mu$
  4. If $X \sim N(\mu, \sigma^2)$, then $Z = \frac{X - \mu}{\sigma} \Rightarrow Z \sim N(0, 1)$
  5. If $X \sim N(\mu_1, \sigma_1^2)$ and $Y \sim N(\mu_2, \sigma_2^2)$, then $Z = X + Y$ is also normally distributed:
     - $Z \sim N(\mu_1 + \mu_2, \sigma_1^2 + \sigma_2^2)$

- We denote the **PDF of the standard normal** by $\phi(x)$.
- The **CDF** is $\Phi(x) = P(X \leq x)$.

- **Example:** Blood Pressure is $N(80, 144)$
  - Mild hypertension is $90 \leq DBP \leq 100$
  - During a random day, what is $P(\text{mild hypertensive})$?

  - Solution:
    - $P(90 \leq X \leq 100)$
    - $= P(X \leq 100) - P(X \leq 90)$
    - $= \text{pnorm}(100, \text{mean} = 80, \text{sd} = \sqrt{144}) - \text{pnorm}(90, \text{mean} = 80, \text{sd} = \sqrt{144})$
    - $\approx 0.1545$

    - XYZ Finish this figure:
  ```{r}
  #| echo: false
  mean_bp <- 80
  sd_bp <- sqrt(144)
  p_100 <- pnorm(100, mean = mean_bp, sd = sd_bp)
  p_90 <- pnorm(90, mean = mean_bp, sd = sd_bp)
  prob_mild_hypertension <- p_100 - p_90
  x_values <- seq(50, 110, by = 0.1)
  y_values <- dnorm(x_values, mean = mean_bp, sd = sd_bp)
  data <- data.frame(x = x_values, y = y_values)
  ggplot(data, aes(x = x, y = y)) +
    geom_line(size = 1) +
    geom_area(data = subset(data, x >= 90 & x <= 100), fill = "lightblue", alpha = 0.5) +
    geom_vline(xintercept = c(90, 100), linetype = "dashed") +
    labs(
      x = "Blood Pressure (DBP)",
      y = "Density",
      title = "Normal Distribution of Blood Pressure with Mild Hypertension Range"
    ) +
    theme_minimal()
  ```

- **Example:** Tree diameter $\sim N(8, 2^2)$ (in inches)
  - What is the probability that a tree has a diameter > 12 inches?
  - **Solution:** $1 - \text{pnorm}(12, 8, 2) \approx 0.02275$

- If $X_1, \dots, X_n$ are random variables and
  $$ L = \sum_{i=1}^n c_i X_i $$
  for $c_i$ constants (not random variables), then
  - $E(L) = \sum_{i=1}^n c_i E(X_i)$
  - $\text{Var}(L) = \sum_{i=1}^n c_i^2 \text{Var}(X_i)$

- If the $X_i$ are also normally distributed, then
  $L \sim N(E(L), \text{Var}(L))$

- **Example:** 
  - $X =$ serum creatinine level for Caucasian individual
  - $Y =$ serum creatinine level for Black individual
  - $X \sim N(1.3, 0.25)$
  - $Y \sim N(1.5, 0.25)$
  - What is the distribution of the average level for one Caucasian and one Black individual chosen at random?

  - Let $Z = \frac{1}{2} X + \frac{1}{2} Y$
  - Then $Z \sim N(1.4, 0.125)$

  - **Calculations**:
    - $E(Z) = \frac{1}{2}(1.3) + \frac{1}{2}(1.5) = 1.4$
    - $\text{Var}(Z) = \frac{1}{4}(0.25) + \frac{1}{4}(0.25) = 0.125$


- **Normal Approximation to Binomial** (rule of thumb):
  - If $X \sim \text{Bin}(n, p)$ and $np(1 - p) \geq 5$, then
    $X \approx N(np, np(1 - p))$

- Let $X \sim \text{Bin}(n, p)$, $Y \sim N(np, np(1 - p))$.
  - Then $P(a \leq X \leq b) \approx P\left(a - \frac{1}{2} \leq X \leq b + \frac{1}{2}\right)$ (continuity correction).

- We will use this for **2-sample binomial tests**.

- **Why?** Let $T_1, T_2, \dots, T_n$ be $n$ independent Bernoulli trials.
  - $T_i = \begin{cases} 1 & \text{with probability } p \\ 0 & \text{with probability } 1 - p \end{cases}$
  - $X = T_1 + T_2 + \dots + T_n = \sum_{i=1}^n T_i$

- **Central Limit Theorem** says normal for large $n$.

- **Normal Approximation to Poisson**:
  - If $X \sim \text{Pois}(\mu)$, then $Y \sim N(\mu, \mu)$.
  - Rule of thumb: For $\mu \geq 10$, 
    $P(a \leq X \leq b) \approx P(a - 1 \leq Y \leq b + 1)$

- **Exercise 5.12 – 5.13 of Rosner**:
  - Of men 30-34 who have smoked:
    - $X = $ # years a man has smoked
    - $Y = $ # of years smoked by women in age group
    - $X \sim N(12.8, 5.1^2)$
    - $Y \sim N(9.3, 3.2^2)$

  - **Q1:** What proportion of men have smoked for more than 20 years? Women?

    - Men: 
      $1 - \text{pnorm}(20, \text{mean} = 12.8, \text{sd} = 5.1) \approx 0.0794$
      
  ```{r}
  #| echo: false
  mean_men <- 12.8
  sd_men <- 5.1
  mean_women <- 9.3
  sd_women <- 3.2
  
  prob_men <- 1 - pnorm(20, mean = mean_men, sd = sd_men)
  prob_women <- 1 - pnorm(20, mean = mean_women, sd = sd_women)
  
  x_values_men <- seq(0, 30, by = 0.1)
  y_values_men <- dnorm(x_values_men, mean = mean_men, sd = sd_men)
  data_men <- data.frame(x = x_values_men, y = y_values_men)
  
  x_values_women <- seq(0, 30, by = 0.1)
  y_values_women <- dnorm(x_values_women, mean = mean_women, sd = sd_women)
  data_women <- data.frame(x = x_values_women, y = y_values_women)
  
  ggplot(data_men, aes(x = x, y = y)) +
    geom_line(size = 1) +
    geom_area(data = subset(data_men, x > 20), fill = "lightblue", alpha = 0.5) +
    geom_vline(xintercept = 20, linetype = "dashed") +
    labs(
      x = "Years Smoked",
      y = "Density",
      title = "Distribution of Years Smoked (Men, Age 30-34)"
    ) +
    theme_minimal()
  ```
    - Women:
      $1 - \text{pnorm}(20, \text{mean} = 9.3, \text{sd} = 3.2) \approx 0.0041$
  ```{r}
  #| echo: false
  ggplot(data_women, aes(x = x, y = y)) +
    geom_line(size = 1) +
    geom_area(data = subset(data_women, x > 20), fill = "lightgreen", alpha = 0.5) +
    geom_vline(xintercept = 20, linetype = "dashed") +
    labs(
      x = "Years Smoked",
      y = "Density",
      title = "Distribution of Years Smoked (Women, Age 30-34)"
    ) +
    theme_minimal()
  ```

- **Exercise 5.12 – 5.13**: 
  - The Christmas Bird Count is a holiday tradition in a boring part of Massachusetts.

  - **Data**:
    - Year and Number of Birds ($x_i$):
      - 2006: 76
      - 2007: 47
      - 2008: 63
      - 2009: 53
      - 2010: 62
      - 2011: 69
      - 2012: 62

    - $\sum x_i = 432$
    - $\sum x_i^2 = 27,717$

  - **Questions**:
    1. What is the mean number of birds?
       - $$ \bar{x} = \frac{\sum x_i}{n} = \frac{432}{7} \approx 61.71 $$

    2. What is the standard deviation?
       - Calculate the variance:
         $$ \text{Variance} = \frac{\sum x_i^2}{n} - \left(\frac{\sum x_i}{n}\right)^2 $$
         $$ = \frac{27,717}{7} - \left(\frac{432}{7}\right)^2 $$
         $$ \approx 78.78 $$
       - Standard deviation:
         $$ \text{SD} = \sqrt{\text{Variance}} = \sqrt{78.78} \approx 8.876 $$

    3. Suppose the number of birds is normally distributed with the same mean and SD as parts 1 and 2. What is the probability of at least 60 birds? Apply the continuity correction.
       - $$ 1 - \text{pnorm}(59.5, \text{mean} = 61.71, \text{sd} = 8.876) $$
       - $$ \approx 0.5983 $$
    
    4. Find the "normal range" $(L, U)$ (integers) such that:
       - $L$ = 15th percentile
       - $U$ = 85th percentile
       - $$ \text{qnorm}(c(0.15, 0.85), \text{mean} = 61.71, \text{sd} = 8.876) $$
       - $$ \approx (52.51, 70.91) $$
       - Thus, 52 to 70 is the "normal range."
    
    5. What is the probability of having a count $\geq U$ at least once during a 10-year period?
       - $$ P(X \geq U) \approx 1 - \text{pnorm}(79.5, \text{mean} = 61.71, \text{sd} = 8.876) $$
       - $$ \approx 0.02252 $$
    
       - Let $Y = $ # of years $\geq U \sim \text{Bin}(10, 0.02252)$.
       - $$ P(Y \geq 1) = 1 - P(Y = 0) = 1 - \text{dbinom}(0, 10, 0.02252) $$
       - $$ \approx 0.2037 $$

