---
title: "Chapter 5: Continuous Probability Distributions"
author: "David Gerard"
format: html
date: today
---

- A continuous random variable "takes on decimal values."

- For such random variables, the probability at any specific value is 0.

-   Example: 
    - $\Pr(\text{a man is exactly } 6', 2.357921784123'') \approx 0$
    - $\Pr(\text{a man is exactly } 6'2'') \approx 0$
    - Men are a little above or a little below.

-   But, we know some regions are more likely than others.
    - $\Pr(5' \leq X \leq 7') > \Pr(0' \leq X \leq 1')$

- We describe this intuition with a PDF.

- A **Probability Density Function** of a random variable $X$ is a function $f$ such that:
  $$
  \Pr(a \leq X \leq b) = \text{area below curve between } a \text{ and } b
  $$

- The **CDF** is again the $F(x) = \Pr(X \leq x)$.

- Example: $X =$ Serum triglyceride level
  - $\Pr(50 \leq X \leq 100)$
    ```{r}
    #| echo: false
    library(ggplot2)
    df <- data.frame(x = seq(0, 200, length.out = 200))
    df$pdf <- dchisq(x = df$x/8, df = 8)
    df2 <- data.frame(x = seq(50, 100, length.out = 200))
    df2$pdf <- dchisq(x = df2$x/8, df = 8)
    ggplot() +
      geom_line(data = df, mapping = aes(x = x, y = pdf)) +
      theme_minimal() +
      xlab("Serum Triglycerides (mg/dL)") +
      ylab("Density") +
      geom_area(data = df2, mapping = aes(x = x, y = pdf), fill = "blue", alpha = 0.5)
    ```
  - $\Pr(X \leq 100)$
    ```{r}
    #| echo: false
    library(ggplot2)
    df <- data.frame(x = seq(0, 200, length.out = 200))
    df$pdf <- dchisq(x = df$x/8, df = 8)
    df2 <- data.frame(x = seq(0, 100, length.out = 200))
    df2$pdf <- dchisq(x = df2$x/8, df = 8)
    ggplot() +
      geom_line(data = df, mapping = aes(x = x, y = pdf)) +
      theme_minimal() +
      xlab("Serum Triglycerides (mg/dL)") +
      ylab("Density") +
      geom_area(data = df2, mapping = aes(x = x, y = pdf), fill = "blue", alpha = 0.5)
    ```
    
- **Expected value:** $\mu$, average $X$ over many trials.
  $$
  \mu = E[X] = \int_{-\infty}^{\infty} x f(x)\,dx
  $$
  where $f(x)$ = density.

- **Variance:** Average squared distance.
  $$
  \sigma^2 = E\left[(X - \mu)^2\right] = E(X^2) - \mu^2
  $$
  $$
  \sigma^2 = \int_{-\infty}^{\infty} (x - \mu)^2 f(x)\,dx
  $$

- **Most common continuous distribution:** Normal distribution
  - If $X \sim N(\mu, \sigma^2)$, then
  $$
  f(x) = \frac{1}{\sqrt{2\pi \sigma^2}} \exp\left[-\frac{1}{2\sigma^2}(x-\mu)^2\right]
  $$
  - Density depends on $\sigma^2$ (variance) and $\mu$ (mean).

- Also, if $X \sim N(\mu, \sigma^2)$, then $E(X) = \mu$, $\mathrm{Var}(X) = \sigma^2$.

-   Normal distribution plot:
    ```{r}
    #| echo: false
    #| warning: false
    x_values <- seq(-4, 4, by = 0.01)
    y_values <- dnorm(x_values)
    data <- data.frame(
      x = x_values,
      y = y_values
    )
    ggplot(data, aes(x = x, y = y)) +
      geom_line(size = 1) +
      geom_area(aes(x = ifelse(x >= -3 & x <= 3, x, NA)), fill = "lightblue", alpha = 0.5) +
      geom_vline(xintercept = c(-3, -2, -1, 0, 1, 2, 3), linetype = "dashed") +
      labs(
        x = "x",
        y = "Density",
        title = "Normal Distribution with Mean and Standard Deviations"
      ) +
      scale_x_continuous(breaks = c(-3, -2, -1, 0, 1, 2, 3), labels = c(expression(mu - 3*sigma), expression(mu - 2*sigma), expression(mu - sigma), expression(mu), expression(mu + sigma), expression(mu + 2*sigma), expression(mu + 3*sigma))) +
      theme_minimal()
    ```

-   Normal densities with different means.
    ```{r, echo=FALSE}
    library(ggplot2)
    library(ggthemes)
    x <- seq(-10, 10, length = 100)
    y1 <- dnorm(x = x, mean = 0, sd = 2)
    y2 <- dnorm(x = x, mean = -4, sd = 2)
    y3 <- dnorm(x = x, mean = 4, sd = 2)
    dfdat <- data.frame(x = rep(x, 3), 
                        y = c(y1, y2, y3), 
                        z = factor(rep(c(1, 2, 3), each = length(x))))
    
    ggplot(data = dfdat, mapping = aes(x = x, y = y, color = z, lty = z)) +
      geom_line(lwd=1) + 
      ggthemes::scale_color_colorblind() +
      theme(legend.position="none") +
      ylab("f(x)") +
      xlab("x") +
      theme_bw() +
      guides(color = "none", linetype = "none")
    ```
    
-   Normal densities with different variances
    ```{r, echo=FALSE}
    x <- seq(-10, 10, length = 100)
    y1 <- dnorm(x = x, mean = 0, sd = 1)
    y2 <- dnorm(x = x, mean = 0, sd = 2)
    y3 <- dnorm(x = x, mean = 0, sd = 4)
    dfdat <- data.frame(x = rep(x, 3), 
                        y = c(y1, y2, y3), 
                        z = factor(rep(c(1, 2, 3), each = length(x))))
    
    ggplot(data = dfdat, mapping = aes(x = x, y = y, color = z, lty = z)) +
      geom_line(lwd=1) + 
      ggthemes::scale_color_colorblind() +
      theme(legend.position="none") +
      ylab("f(x)") +
      xlab("x") +
      theme_bw() +
      guides(color = "none", linetype = "none")
    ```

- The **standard normal distribution** is $N(0,1)$.

- **Properties:**
  1. 68–95–99.7 rule:
     - 68% of area within $\pm 1\sigma$
     - 95% of area within $\pm 2\sigma$
     - 99.7% of area within $\pm 3\sigma$
  2. Symmetric: $f(\mu - x) = f(\mu + x)$
  3. $\mu =$ median
  4. If $X \sim N(\mu, \sigma^2)$, then $Z = \frac{X - \mu}{\sigma} \sim N(0, 1)$
  5. If $X \sim N(\mu_1, \sigma_1^2)$ and $Y \sim N(\mu_2, \sigma_2^2)$ are independent, then
     $$
     Z = X + Y \sim N(\mu_1 + \mu_2, \sigma_1^2 + \sigma_2^2)
     $$

- We denote PDF of standard normal by $\phi(x)$

- The CDF is $\Phi(x) = \Pr(X \leq x)$

- **Example:**
  - Blood Pressure $\sim N(80, \sqrt{144})$
  - Mild hypertension is $90 \leq \text{DBP} \leq 100$
    - Units are in mmHg
  - Individuals are randomly sampled
  - What is $\Pr(\text{mild hypertensive})$?

  \begin{align*}
    \Pr(90 \leq X \leq 100) &= \Pr(X \leq 100) - \Pr(X < 90)\\
    &= \texttt{pnorm(100, mean = 80, sd = sqrt(144))} - \texttt{pnorm(90, mean = 80, sd = sqrt(144))}\\
    &= 0.1545
  \end{align*}

- Visualizations of normal distribution regions **XYZ IMAGE HERE**

- **Exercise:** Tree diameter $\sim N(8, 2^2)$ (in inches)  
  What is the probability that the tree has diameter $> 12$ in?

  **Solution:**  
  $$
  1 - \texttt{pnorm}(12, 8, 2) = 0.02275
  $$

---

- If $X_1, \dots, X_n$ are random variables and  
  $$
  L = \sum_{i=1}^n c_i X_i \quad \text{for } c_i \text{ constants (not r.v.s)}
  $$
  then  
  $$
  \mathbb{E}[L] = \sum_{i=1}^n c_i \mathbb{E}[X_i]
  $$
  $$
  \mathrm{Var}(L) = \sum_{i=1}^n c_i^2 \mathrm{Var}(X_i)
  $$

- If the $X_i$ are also normally distributed, then  
  $$
  L \sim N(\mathbb{E}[L], \mathrm{Var}(L))
  $$

- **Example:**  
  Let $X$ = serum creatinine level for Caucasian individual  
  Let $Y$ = serum creatinine level for Black individual  

  $$
  X \sim N(1.3, 0.125), \quad Y \sim N(1.5, 0.25)
  $$

  What is the distribution of the **average** level for one Caucasian and one Black individual chosen at random?

  Let
  $$
  Z = \frac{1}{2}X + \frac{1}{2}Y \Rightarrow Z \sim N(1.4, 0.175)
  $$

  - $\mathbb{E}(Z) = \frac{1}{2}(1.3 + 1.5) = 1.4$
  - $\mathrm{Var}(Z) = \frac{1}{4}(0.125 + 0.25) = 0.175$

- **Normal Approximation to Binomial** (rule of thumb):

  If $X \sim \mathrm{Bin}(n, p)$ and $np(1-p) \geq 5$, then
  $$
  X \approx N(np, np(1-p))
  $$

- Let $X \sim \mathrm{Bin}(n, p)$, and let $Y \sim N(np, np(1 - p))$.

  Then with **continuity correction**:
  $$
  \Pr(a \leq X \leq b) \approx \Pr\left(a - \frac{1}{2} \leq X \leq b + \frac{1}{2}\right)
  $$

- We will use this for 2-sample binomial tests.

- **Why?**  
  Let $T_1, T_2, \dots, T_n$ be $n$ independent Bernoulli trials:
  $$
  T_i =
  \begin{cases}
  1 & \text{w.p. } p \\
  0 & \text{w.p. } 1 - p
  \end{cases}
  $$

  Let
  $$
  X = T_1 + T_2 + \dots + T_n = \sum T_i
  $$

- The **Central Limit Theorem** says normal for large $n$.

- **Normal Approximation to Poisson**

  If $X \sim \mathrm{Poisson}(\mu)$, then  
  $Y \sim N(\mu, \mu)$

  - Rule of thumb: for $\mu \geq 10$  
  $$
  \Pr(a \leq X \leq b) \approx \Pr\left(a - \frac{1}{2} \leq Y \leq b + \frac{1}{2}\right)
  $$

---

- **Exercise 5.12 – 5.13 of Rosner**

  - Of men aged 30–34 who have smoked:
    - $X$ = number of years a man has smoked
    - $Y$ = number of years smoked by women in age group

  $$
  X \sim N(12.8, 5.1^2), \quad Y \sim N(9.3, 3.2^2)
  $$

- **Q1:** What proportion of men have smoked for more than 20 years? Women?

  - Men:
    $$
    1 - \texttt{pnorm}(20, \text{mean} = 12.8, \text{sd} = 5.1) = 0.07901
    $$

  - Women:
    $$
    1 - \texttt{pnorm}(20, \text{mean} = 9.3, \text{sd} = 3.2) = 0.004133
    $$

- **XYZ IMAGE HERE**

- **Exercise 5.126 – 5.130**

  Christmas Bird Count is a holiday tradition in a boring part of Massachusetts.

  | Year | $x_i$ |
  |------|-------|
  | 2005 | 76    |
  | 2006 | 47    |
  | 2007 | 63    |
  | 2008 | 53    |
  | 2009 | 62    |
  | 2010 | 64    |
  | 2011 | 67    |

  $$
  \sum x_i = 432, \quad \sum x_i^2 = 27,\!717
  $$

1. What is the **mean** number of birds?

   $$
   \bar{x} = \frac{432}{7} = 61.71
   $$

2. What is the **standard deviation**?

   $$
   \frac{1}{7}(27717) - \left(\frac{432}{7}\right)^2 = 78.78
   $$

   $$
   \mathrm{SD} = \sqrt{78.78} = 8.876
   $$

3. Suppose number of birds is normal with same mean and SD as previous years.  
   What is the probability of at least 60 birds? Apply continuity correction.

   $$
   1 - \texttt{pnorm}(59.5, \text{mean} = 61.71, \text{sd} = 8.876) = \boxed{0.5983}
   $$

4. Find **"normal range"** $(L, U)$ (integers) such that:
   - $L$ = 5th percentile  
   - $U$ = 95th percentile

   $$
   \texttt{qnorm}(c(0.05, 0.95), \text{mean} = 61.71, \text{sd} = 8.876) = (52.51, 70.91)
   $$

   - So the normal range is **52 to 80**

5. What is the probability that $X \geq U$ at least once during a 10-year period?

   $$
   \Pr(X \geq U) = 1 - \texttt{pnorm}(79.5, \text{mean} = 61.71, \text{sd} = 8.876) = 0.02252
   $$

   Let $Y$ = number of years $\geq U$  
   $Y \sim \mathrm{Bin}(10, 0.02252)$

   $$
   \Pr(Y \geq 1) = 1 - \Pr(Y = 0) = 1 - \texttt{dbinom}(0, 10, 0.02252) = \boxed{0.2037}
   $$
