---
title: "Chapter 5 Notes: Continuous Distributions"
author: "David Gerard"
format: html
date: today
---

- A continuous random variable "takes on decimal values."
- For such random variables, the probability at any specific value is $0$.

- **Example:** 
  - What is $P(\text{a man is exactly } 6', 2.35792471613")$?
  - $P(\text{a man is exactly } 6'2") \approx 0$
  - Men are a little above or a little below.

- However, we know some regions are more likely than others.
  - $P(5' \leq X \leq 7') > P(0 \leq X \leq 1')$

- We describe this intuition with a **PDF** (Probability Density Function).
- A Probability Density Function of a random variable $X$ is a function, $f$, where:
  - $$ P(a \leq X \leq b) = \text{area below curve between } a \text{ and } b $$

- The CDF (Cumulative Distribution Function) is again the $F(x) = P(X \leq x)$.

- **Example:** Let $X = \text{Serum triglyceride level}$

  ```{r}
  #| echo: false
  library(ggplot2)
  df <- data.frame(x = seq(0, 200, length.out = 200))
  df$pdf <- dchisq(x = df$x/8, df = 8)
  df2 <- data.frame(x = seq(50, 100, length.out = 200))
  df2$pdf <- dchisq(x = df2$x/8, df = 8)
  ggplot() +
    geom_line(data = df, mapping = aes(x = x, y = pdf)) +
    theme_minimal() +
    xlab("Serum Triglycerides (mg/dL)") +
    ylab("Density") +
    geom_area(data = df2, mapping = aes(x = x, y = pdf), fill = "blue", alpha = 0.5) +
    ggtitle("Pr(50 ≤ X ≤ 100)")
  ```

  ```{r}
  #| echo: false
  library(ggplot2)
  df <- data.frame(x = seq(0, 200, length.out = 200))
  df$pdf <- dchisq(x = df$x/8, df = 8)
  df2 <- data.frame(x = seq(0, 100, length.out = 200))
  df2$pdf <- dchisq(x = df2$x/8, df = 8)
  ggplot() +
    geom_line(data = df, mapping = aes(x = x, y = pdf)) +
    theme_minimal() +
    xlab("Serum Triglycerides (mg/dL)") +
    ylab("Density") +
    geom_area(data = df2, mapping = aes(x = x, y = pdf), fill = "blue", alpha = 0.5) +
    ggtitle("Pr(X ≤ 100)")
  ```
